---
title: "Bayesian Unobserved Component Models"
author:
  - name: Tomasz Woźniak
    url: https://github.com/donotdespair
    orcid: 0000-0003-2212-2378
  - name: Yansong Yang
    url: https://github.com/russellyang728
  
    execute:
  #echo: false
citation: 
  issued: 2024-05-01
  url: https://donotdespair.github.io/Bayesian-Unobserved-Component-Models/
  doi: 10.26188/25814617
bibliography: references.bib
---

> **Abstract.** We present the basics of Bayesian estimation and inference for unobserved component models on the example of a local-level model. 
> The range of topics includes the conjugate prior analysis using normal-inverted-gamma 2 distribution and its extensions focusing on hierarchical modelling, conditional heteroskedasticity, and Student-t error terms. 
> We scrutinise Bayesian forecasting and sampling from the predictive density.
>
> **Keywords.** Unobserved Component Models, Local-Level Model, State-Space Bayesian Inference, Forecasting, Heteroskedasticity, Hierarchical Modelling, Gibbs Sampler, Simulation Smoother, Precision Sampling

# Unobserved component models

Unobserved Component (UC) models are a popular class of models in macroeconometrics that use the state-space representation for unit-root nonstationary time series. 
The simple formulation of the model equations decomposing the series into a non-stationary and stationary component facilitates economic interpretations and good forecasting performance.

# A simple local-level model

The model is set for a univariate time series whose observation at time
$t$ is denoted by $y_t$. 
It decomposes the variable into a stochastic trend component, $\tau_t$, and a stationary error component, $\epsilon_t$. 
The former follows a Gaussian random walk process with the conditional variance $\sigma_\eta^2$, and the latter is zero-mean normally distributed with the variance $\sigma^2$.
These are expressed as the model equations:
\begin{align}
y_t &= \tau_t + \epsilon_t,\\
\tau_t &= \tau_{t-1} + \eta_t,\\
\epsilon_t &\sim\mathcal{N}\left(0, \sigma^2\right),\\
\eta_t &\sim\mathcal{N}\left(0, \sigma_\eta^2\right),
\end{align}
where the initial condition $\tau_0$ is a parameter of the model.

## Matrix notation for the model

To simplify the notation and the derivations introduce matrix notation for the model. Let $T$ be the available sample size for the variable $y$. 
Define a $T$-vector of zeros, $\mathbf{0}_T$, and of ones, $\boldsymbol\imath_T$, the identity matrix of order $T$, $\mathbf{I}_T$, as well as $T\times1$ vectors:
\begin{align}
\mathbf{y} = \begin{bmatrix} y_1\\ \vdots\\ y_T \end{bmatrix},\quad
\boldsymbol\tau = \begin{bmatrix} \tau_1\\ \vdots\\ \tau_T \end{bmatrix},\quad
\boldsymbol\epsilon = \begin{bmatrix} \epsilon_1\\ \vdots\\ \epsilon_T \end{bmatrix},\quad
\boldsymbol\eta = \begin{bmatrix} \eta_1\\ \vdots\\ \eta_T \end{bmatrix},\qquad
\mathbf{i} = \begin{bmatrix} 1\\0\\ \vdots\\ 0 \end{bmatrix},
\end{align}
and a $T\times T$ matrix $\mathbf{H}$ with the elements:
\begin{align}
\mathbf{H} = \begin{bmatrix}
1 & 0 & \cdots & 0 & 0\\
-1 & 1 & \cdots & 0 & 0\\
0 & -1 & \cdots & 0 & 0\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
0 & 0 & \cdots & 1 & 0\\
0 & 0 & \cdots & -1 & 1
\end{bmatrix}.
\end{align}

Then the model can be written in a concise notation as:
\begin{align}
\mathbf{y} &= \mathbf{\tau} + \boldsymbol\epsilon,\\
\mathbf{H}\boldsymbol\tau &= \mathbf{i} \tau_0 + \boldsymbol\eta,\\
\boldsymbol\epsilon &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma^2\mathbf{I}_T\right),\\
\boldsymbol\eta &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma_\eta^2\mathbf{I}_T\right).
\end{align}

## Likelihood function

The model equations imply the predictive density of the data vector
$\mathbf{y}$. To see this, consider the model equation as a linear
transformation of a normal vector $\boldsymbol\epsilon$. Therefore, the data vector follows a multivariate normal distribution given by:
\begin{align}
\mathbf{y}\mid \boldsymbol\tau, \sigma^2 &\sim\mathcal{N}_T\left(\boldsymbol\tau, \sigma^2\mathbf{I}_T\right).
\end{align}

This distribution determines the shape of the likelihood function that
is defined as the sampling data density: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y})\equiv p\left(\mathbf{y}\mid\boldsymbol\tau, \sigma^2 \right).
\end{align}

The likelihood function that for the sake of the estimation of the parameters, and after plugging in data in place of $\mathbf{y}$, is considered a function of parameters $\boldsymbol\tau$ and $\sigma^2$ is given by: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y}) = 
(2\pi)^{-\frac{T}{2}}\left(\sigma^2\right)^{-\frac{T}{2}}\exp\left\{-\frac{1}{2}\frac{1}{\sigma^2}(\mathbf{y} - \boldsymbol\tau)'(\mathbf{y} - \boldsymbol\tau)\right\}.
\end{align}

## Prior distributions

# Bayesian estimation

## Gibbs sampler

## Simulation smoother and precision sampler

## Analytical solution for a joint posterior

# Hierarchical modeling

## Estimating gamma error term variance prior scale


## Estimating inverted-gamma 2 error term variance prior scale

## Estimating the initial condition prior scale

In the univariate local-level model, the initial state $\tau_0$ and the
variance $s^2$ are key parameters determining the behavior of the
stochastic process. For the initial state $\tau_0$, a prior is specified
as an inverse gamma distribution, denoted by
$\tau_0 \sim \text{N}(\underline\tau_0, s^2)$, reflecting our belief
about the distribution of $\tau_0$ before observing the data. While, for
the variance $s^2$, which controls the fluctuation of the process, we
use inverse gamma 2 distribution,
$s^2 \sim \text{IG2}(\underline{s}, \underline{\nu})$.

The full posterior of $s^2$ is derived below: 

$$
\begin{gather*}
&p(s^2|y,\sigma,\tau,\tau_0,\sigma_{\eta}, \underline{s}, \underline{\nu}) \propto
p(y|\tau,\sigma)p(\tau|\tau_0,\sigma_{\eta}^2)p(\tau_0|s^2)p(s^2) \\ \\
&\propto p(\tau_0|s^2)p(s^2|\underline{s}, \underline{\nu}) \\ \\
&= (s^2)^{-\frac{1}{2}}\exp\left\{-\frac{1}{2}(\tau_0-\underline{\tau_0})'(s^2)^{-1}(\tau_0-\underline{\tau_0})\right\}(s^2)^{-\frac{\underline{\nu}+2}{2}}\exp\left\{-\frac{1}{2}\frac{\underline{s}}{s^2}\right\} \\ \\
&= (s^2)^{-\frac{1+\underline{\nu}+2}{2}}
\exp \left\{-\frac{1}{2}\frac{(\tau_0-\underline{\tau_0})^2+\underline{s}}{s^2}\right\}
\end{gather*}
$$

Hence, full conditional posterior of $s^2$ in a form of inverse
gamma 2 distribution, where

$$
\begin{gather*}
\overline{\nu} = 1+\underline{\nu} \\
\overline{s} = (\tau_0-\underline{\tau_0})^2+\underline{s}
\end{gather*}
$$
This R code below implements a Bayesian posterior sampling process. The code gives hyper parameters initial value for the prior distribution and then updates the posterior distribution parameters by combining them with data observations.

```{r echo=TRUE}
# Define the prior hyper-parameters
s_prior = 10
nu_prior = 1
tau_0 = 1    #Initial state parameter
tau_0_prior = 1

s2_sampler <- function(total_s, s_prior,tau_0_prior,nu_prior) {
    # Initialize vector to store samples of s2
    s2_posterior <- numeric(total_s)
    
    for (s in 1:total_s) {
      # Update parameters for s2 posterior based on your derivation
      nu_posterior <- nu_prior + 1
      s_posterior <- (tau_0 - tau_0_prior)^2 + s_prior
      
      # Draw next period value for s2 from inverse gamma distribution
      s2_posterior[s] <- s_posterior / rchisq(1, nu_posterior)
      if (s < total_s) {
        s_posterior = s2_posterior[s]
      }
    }
    
    # Prepare the output
    s2_draw <- list(s2_posterior = s2_posterior)
    return(s2_draw)
}
```

The plot visualizes the posterior samples of the $s^2$ parameter over time, derived from the Bayesian updating process.

```{r echo=FALSE}
qq = s2_sampler(100, s_prior,tau_0_prior,nu_prior)
plot.ts(qq$s2_posterior, ylab = "s2_posterior")
```


## Student-t prior for the trend component

## Estimating Student-t degrees of freedom parameter

## Laplace prior for the trend component

# Model extensions

## Estimation of autoregressive parameters for cycle component

## Autoregressive cycle component

## Random walk with  time-varying drift parameter

## Student-t error terms

## Conditional heteroskedasticity

# Bayesian forecasting

## Predictive density

## Sampling from the predictive density




## References {.unnumbered}
