---
title: "Bayesian Unobserved Component Models"
author:
  - name: Tomasz Woźniak
    url: https://github.com/donotdespair
    orcid: 0000-0003-2212-2378
  - name: Ben Gussen
    url: https://github.com/bgussen
    orcid: 0000-0003-4406-4076
  - name: Rui Liu
    url: https://github.com/rruiliu1005
    orcid: 0009-0008-9348-8581
  - name: Zheyuan Li
    url: https://github.com/lzyzero3
  
    execute:
  #echo: false
citation: 
  issued: 2024-05-01
  url: https://donotdespair.github.io/Bayesian-Unobserved-Component-Models/
  doi: 10.26188/25814617
bibliography: references.bib
---

> **Abstract.** We present the basics of Bayesian estimation and inference for unobserved component models on the example of a local-level model. 
> The range of topics includes the conjugate prior analysis using normal-inverted-gamma 2 distribution and its extensions focusing on hierarchical modelling, conditional heteroskedasticity, and Student-t error terms. 
> We scrutinise Bayesian forecasting and sampling from the predictive density.
>
> **Keywords.** Unobserved Component Models, Local-Level Model, State-Space Bayesian Inference, Forecasting, Heteroskedasticity, Hierarchical Modelling, Gibbs Sampler, Simulation Smoother, Precision Sampling

# Unobserved component models

Unobserved Component (UC) models are a popular class of models in macroeconometrics that use the state-space representation for unit-root nonstationary time series. 
The simple formulation of the model equations decomposing the series into a non-stationary and stationary component facilitates economic interpretations and good forecasting performance.

# A simple local-level model

The model is set for a univariate time series whose observation at time
$t$ is denoted by $y_t$. 
It decomposes the variable into a stochastic trend component, $\tau_t$, and a stationary error component, $\epsilon_t$. 
The former follows a Gaussian random walk process with the conditional variance $\sigma_\eta^2$, and the latter is zero-mean normally distributed with the variance $\sigma^2$.
These are expressed as the model equations:
\begin{align}
y_t &= \tau_t + \epsilon_t,\\
\tau_t &= \tau_{t-1} + \eta_t,\\
\epsilon_t &\sim\mathcal{N}\left(0, \sigma^2\right),\\
\eta_t &\sim\mathcal{N}\left(0, \sigma_\eta^2\right),
\end{align}
where the initial condition $\tau_0$ is a parameter of the model.

## Matrix notation for the model

To simplify the notation and the derivations introduce matrix notation for the model. Let $T$ be the available sample size for the variable $y$. 
Define a $T$-vector of zeros, $\mathbf{0}_T$, and of ones, $\boldsymbol\imath_T$, the identity matrix of order $T$, $\mathbf{I}_T$, as well as $T\times1$ vectors:
\begin{align}
\mathbf{y} = \begin{bmatrix} y_1\\ \vdots\\ y_T \end{bmatrix},\quad
\boldsymbol\tau = \begin{bmatrix} \tau_1\\ \vdots\\ \tau_T \end{bmatrix},\quad
\boldsymbol\epsilon = \begin{bmatrix} \epsilon_1\\ \vdots\\ \epsilon_T \end{bmatrix},\quad
\boldsymbol\eta = \begin{bmatrix} \eta_1\\ \vdots\\ \eta_T \end{bmatrix},\qquad
\mathbf{i} = \begin{bmatrix} 1\\0\\ \vdots\\ 0 \end{bmatrix},
\end{align}
and a $T\times T$ matrix $\mathbf{H}$ with the elements:
\begin{align}
\mathbf{H} = \begin{bmatrix}
1 & 0 & \cdots & 0 & 0\\
-1 & 1 & \cdots & 0 & 0\\
0 & -1 & \cdots & 0 & 0\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
0 & 0 & \cdots & 1 & 0\\
0 & 0 & \cdots & -1 & 1
\end{bmatrix}.
\end{align}

Then the model can be written in a concise notation as:
\begin{align}
\mathbf{y} &= \mathbf{\tau} + \boldsymbol\epsilon,\\
\mathbf{H}\boldsymbol\tau &= \mathbf{i} \tau_0 + \boldsymbol\eta,\\
\boldsymbol\epsilon &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma^2\mathbf{I}_T\right),\\
\boldsymbol\eta &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma_\eta^2\mathbf{I}_T\right).
\end{align}

## Likelihood function

The model equations imply the predictive density of the data vector
$\mathbf{y}$. To see this, consider the model equation as a linear
transformation of a normal vector $\boldsymbol\epsilon$. Therefore, the data vector follows a multivariate normal distribution given by:
\begin{align}
\mathbf{y}\mid \boldsymbol\tau, \sigma^2 &\sim\mathcal{N}_T\left(\boldsymbol\tau, \sigma^2\mathbf{I}_T\right).
\end{align}

This distribution determines the shape of the likelihood function that
is defined as the sampling data density: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y})\equiv p\left(\mathbf{y}\mid\boldsymbol\tau, \sigma^2 \right).
\end{align}

The likelihood function that for the sake of the estimation of the parameters, and after plugging in data in place of $\mathbf{y}$, is considered a function of parameters $\boldsymbol\tau$ and $\sigma^2$ is given by: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y}) = 
(2\pi)^{-\frac{T}{2}}\left(\sigma^2\right)^{-\frac{T}{2}}\exp\left\{-\frac{1}{2}\frac{1}{\sigma^2}(\mathbf{y} - \boldsymbol\tau)'(\mathbf{y} - \boldsymbol\tau)\right\}.
\end{align}

## Prior distributions

# Bayesian estimation

## Gibbs sampler

## Simulation smoother and precision sampler

## Analytical solution for a joint posterior
$$\underset{T \times 1}{y} = \begin{pmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{pmatrix} \qquad 
\underset{T \times 1}{\tau} = \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\\tau_T \end{pmatrix} \qquad 
\underset{T \times 1}{\epsilon} = \begin{pmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{pmatrix} \qquad 
\underset{T \times 1}{\eta} = \begin{pmatrix} \eta_1 \\ \eta_2 \\ \vdots \\ \eta_T \end{pmatrix}$$

$$\underset{T \times T}{H} = 
\begin{pmatrix} 1 & 0 & 0 & \cdots & 0 & 0 \\ 
-1 & 1 & 0 & \cdots & 0 & 0 \\ 
0 & -1 & 1 & \cdots & 0 & 0 \\ 
\vdots & \vdots & \ddots & \ddots & \vdots & \vdots \\ 
0&0&0 & \cdots & 1 & 0  \\ 
0 & 0 & 0 &0& -1 & 1 \end{pmatrix}$$


In this setting, we have out state equations as: 

$$y = \tau + \epsilon \quad\quad\qquad \qquad \qquad \qquad \text{measurement equation}$$

$$\color{purple}{\mathbf{H\tau = \eta \implies \tau = H^{-1} \eta}} \quad \qquad \color{black}{  \qquad \qquad\text{state equation 1}}$$

The state equations in matrix notation are derived as follows: 

(1) In an unobserved component model, the measurement equation is defined as $y_t = \tau_t + \epsilon_t$, in matrix notation, we have: 

$$y = \begin{pmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{pmatrix} = \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\ \tau_T \end{pmatrix} + \begin{pmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{pmatrix} = \tau + \epsilon$$
(2) We have for each t = 1,...,T, $\tau_t = \tau_{t-1} + \eta_t$, and $\tau_0 = 0$, so in matrix notation: 

$$\tau = \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\ \tau_T \end{pmatrix} = \begin{pmatrix} \eta_1 \\ 
\tau_1+ \eta_2 \\
\tau_2+ \eta_3 \\ 
\vdots \\ 
\tau_{T-1} + \eta_T  \end{pmatrix} = \begin{pmatrix} 0 \\ \tau_1 \\ \vdots \\ \tau_{T-1} \end{pmatrix} + \eta$$

Moving the lagged $\tau$ terms on the right hand side to the left:

$$\begin{pmatrix} 
1 & 0 & \cdots & 0 \\
-1 & 1 & \cdots & 0 \\ 
\vdots & \ddots & \ddots & \vdots \\ 
0 & 0 & -1 & 1
\end{pmatrix} \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\ \tau_T \end{pmatrix}  = \begin{pmatrix} \tau_1 \\ \tau_2-\tau_1 \\ \vdots \\ \tau_T - \tau_{T-1} \end{pmatrix} = \eta$$

Hence, we have: 
$$H \tau = \eta \implies \tau = H^{-1}\eta$$

We assume $\mathbf{\color{purple}{\eta \sim \mathcal{N}(0_T, c\sigma^2I_T)}}$ and $\sigma^2 \sim \mathcal{IG2}(s_{prior}, \nu_{prior})$

The parameters to be estimated are: $\mathbf{\color{purple}{\tau, \sigma^2}}$

### 2.Prior distributions 

(1) **Prior distribution for** $\tau$ 

From the state equation, we have: 
$$\tau =  H^{-1}\eta$$
we know that $$\mathbf{\color{purple}{\eta \sim \mathcal{N}(0_T, c \sigma^2 I_T) \implies H^{-1}\eta \sim \mathcal{N}(0_T, c\sigma^2(H^T H)^{-1})}}$$
since $Var(H^{-1}\eta) = H^{-1}Var(\eta)(H^{-1})^T =c\sigma^2(H^TH)^{-1}$
Then, the prior distribution of $\tau|c\sigma^2$ is 
$$\tau|c\sigma^2 \sim \mathcal{N}(0_T, c\sigma^2(H^TH)^{-1}) \\ \propto (c\sigma^2)^{-\frac{T}{2}}exp\left(-\frac{\tau^TH^T H\tau}{2c\sigma^2}\right)
$$
(2) **Prior assumptions for $\sigma^2$**

$$\sigma^2 \sim \mathcal{IG2}(s_{prior}, \nu_{prior}) \propto (\sigma^2)^\frac{-\nu_{prior}+2}{2}exp\left(-\frac{s_{prior}}{2\sigma^2}\right)$$

### 3. Posterior distributions

1. $\mathbf{p(\tau|y, \alpha, \beta, \sigma^2_e)}$
We have, from the measurement equation: 
$$y = \tau + \epsilon$$
since $\epsilon \sim \mathcal{N}(0_T, \sigma^2I_T)$
$$y|\tau, \sigma^2 \sim \mathcal{N}(\tau, \sigma^2I_T)\propto (\sigma^2)^{-\frac{T}{2}}exp\left(-\frac{1}{2\sigma^2}(y-\tau)^T(y-\tau)\right)$$
and the prior of $\tau$ is: 

$$\tau|c\sigma^2 \sim \mathcal{N}(0_T, c\sigma^2I_T) \\ \propto (\sigma^2)^{-\frac{T}{2}}exp\left(-\frac{\tau^TH^T H\tau}{2c\sigma^2}\right)$$

the prior of $\sigma^2$ is: 
$$p(\sigma^2) \propto (\sigma^2)^-{\frac{\nu_{prior}+2}{2}}exp(-\frac{s_{prior}}{2\sigma^2})$$

The joint posterior of $\tau$ and $\sigma^2$ can be derived as follows: 

$$
p(\tau, \sigma^2 |y) = \frac{p(\tau, \sigma^2, y)}{p(y)} \propto p(\tau, \sigma^2, y) = p(y | \tau, \sigma^2)p(\tau, \sigma^2) =p(y | \tau, \sigma^2)p(\tau|\sigma^2)p(\sigma^2)
$$

$$
\propto (\sigma^2)^{-\frac{T}{2}}exp\left(-\frac{(y-\tau)^T(y-\tau)}{2\sigma^2}\right) \times 
(\sigma^2)^{-\frac{T}{2}}exp\left(-\frac{\tau^TH^T H\tau}{2c\sigma^2}\right) \times
(\sigma^2)^{-\frac{\nu_{prior}+2}{2}} exp(-\frac{s_{prior}}{2\sigma^2})
$$
$$
\propto exp(-\frac{y^Ty - 2\tau^T y + \tau^T\tau + c^{-1}\tau^TH^T H\tau}{2\sigma^2}) \times \\ 
exp(-\frac{s_{prior}}{2\sigma^2})\times(\sigma^2)^{-\frac{2T+\nu_{prior}+2}{2}}
$$
$$ = exp(-\frac{\tau^T(c^{-1}H^T H + I_T)\tau - 2\tau^Ty}{2\sigma^2}) \times \\ 
exp(-\frac{y^Ty + s_{prior}}{2\sigma^2})\times(\sigma^2)^{-\frac{2T+\nu_{prior}+2}{2}} $$

Let $\bar{\Sigma} = (c^{-1}H^T H + I_T)^{-1}$, we have 
$$p(\tau, \sigma^2 |y) \propto exp(-\frac{\tau^T\bar{\Sigma}^{-1}\tau - 2\tau^Ty + y^T\bar{\Sigma}y}{2\sigma^2})exp(-\frac{y^Ty + s_{prior}-y^T\bar{\Sigma}y}{2\sigma^2})\times(\sigma^2)^{-\frac{2T+\nu_{prior}+2}{2}}$$
Hence, $\tau, \Sigma |y \sim \mathcal{NIG2}(\bar{\tau}, \bar{\Sigma}, \bar{\nu}, \bar{s})$ where 
$$\bar{\Sigma} = (c^{-1}H^T H + I_T)^{-1}$$
$$\bar{\tau} = \bar{\Sigma}y$$
$$\bar{\nu} = 2T+\nu_{prior}$$
$$\bar{s} = s_{prior}+y^Ty-y^T\bar{\Sigma}y$$
```{r}
UC.local.simple.Gibbs.sampler    = function(S, starting.values, priors){
  # Estimation of a local level UC model with \tau_0 = 0 and no AR trend. 
  aux     = starting.values
  T       = nrow(aux$Y)
  posteriors    = list(
    tau     = matrix(NA,T,S),
    sigma   = matrix(NA,2,S),
    non.stationary.iterations = rep(NA,S)
  )

  for (s in 1:S){



    # Sampling tau
    ###########################
    V.tau.inv     = priors$c^(-1)*t(aux$H)%*%(aux$H) + diag(T)
    V.tau.inv     = 0.5*(V.tau.inv + t(V.tau.inv))
    b.tau         = aux$Y
    precision.L   = t(bandchol(V.tau.inv))
    epsilon       = rnorm(T)
    b.tau.tmp     = forwardsolve(precision.L, b.tau)
    tau.draw      = backsolve(t(precision.L), b.tau.tmp + epsilon)
    aux$tau       = tau.draw
    
    # Sampling sigma
    ###########################
    sigma.s   = priors$sigma.s +t(aux$Y)%*%aux$Y - t(aux$Y)%*%aux$tau%*%aux$Y
    sigma.nu  = priors$sigma.nu + 2*T
    sigma.draw= sigma.s/rchisq(1,sigma.nu)

    posteriors$tau[,s]     = aux$tau
    posteriors$sigma[,s]   = aux$sigma
    # posteriors$non.stationary.iterations[s] = ns.i
    if (s%%1000==0){cat(" ",s)}
  }

  output      = list(
    posterior = posteriors,
    last.draw = aux
  )
  return(output)
}

```
# Hierarchical modeling

## Estimating gamma error term variance prior scale
To estimate the gamma error term variance, we need to firstly put a prior on the prior scale of and present its full conditional posterior distribution.

The prior is below:

```{=tex}
\begin{align*}
\sigma^2 \mid s &\sim \text{IG2}(s, \nu) \\
s &\sim \mathcal{G}(s, a)
\end{align*}
```


Then we use Gibbs sampling to draw samples from the full conditional distributions: 
```{r}
#| echo: true
#| message: false
#| warning: false
library(MCMCpack)
# Define prior parameters
s_prior <- 2
nu_prior <- 2
a_prior <- 2

# Define the function for the conditional posterior distribution
posterior_sampler <- function(y, n_iter = 1000) {
  # Store samples
  s <- rgamma(1, shape = a_prior, rate = s_prior)
  sigma2 <- rinvgamma(1, shape = s, scale = nu_prior)  

  s_samples <- numeric(n_iter)
  sigma2_samples <- numeric(n_iter)

  for (i in 1:n_iter) {
    shape_sigma2 <- (length(y) / 2) + s
    rate_sigma2 <- (sum((y - mean(y))^2) / 2) + nu_prior
    sigma2 <- rinvgamma(1, shape = shape_sigma2, scale = rate_sigma2)  

    shape_s <- a_prior + 1
    rate_s <- s_prior + sigma2
    s <- rgamma(1, shape = shape_s, rate = rate_s)

    sigma2_samples[i] <- sigma2
    s_samples[i] <- s
  }

  return(list(sigma2_samples = sigma2_samples, s_samples = s_samples))
}

set.seed(123)
y <- rnorm(100, mean = 0, sd = 1)

# Run the sampler
posterior_samples <- posterior_sampler(y, n_iter = 1000)

# Plot posterior distributions
par(mfrow = c(2, 1))
hist(posterior_samples$sigma2_samples, main = expression(paste("Posterior of ", sigma^2)), xlab = expression(sigma^2))
hist(posterior_samples$s_samples, main = "Posterior of s", xlab = "s")
```

Based on the sampling results, we summarize the posterior distribution of these parameters.

As can be seen from the histogram, the values of $\sigma^2$ are mostly concentrated between 0.6 and 1.0. The distribution has a certain skewness, with a small number of higher values on the right, but it is generally concentrated.

Most of the values of $s$ are concentrated between 0 and 2, indicating that the probability of s is higher in this range and the distribution has a longer right tail.

To derive the full conditional posteriors for $\sigma^2$ and $s$, we start with the joint posterior distribution of these parameters given the data $y$.

The joint posterior distribution is below:

```{=tex}
\begin{equation}
p(\sigma^2, s \mid y) \propto p(y \mid \sigma^2) p(\sigma^2 \mid s) p(s)
\end{equation}
```


Full Conditional Posterior for $\sigma^2$ is: 

```{=tex}
\begin{equation}
p(\sigma^2 \mid y, s) \propto (\sigma^2)^{-\left( \frac{T}{2} + s + 1 \right)} \exp \left( -\frac{1}{\sigma^2} \left( \frac{(y - \tau)'(y - \tau)}{2} + \nu \right) \right)
\end{equation}
```


Full Conditional Posterior for $s$ is:

```{=tex}
\begin{equation}
p(s \mid \sigma^2) \propto s^{a-1} \nu^s (\sigma^2)^{-(s+1)} \exp \left( -s - \frac{\nu}{\sigma^2} \right)
\end{equation}
```


## Estimating inverted-gamma 2 error term variance prior scale

## Estimating the initial condition prior scale

## Student-t prior for the trend component

## Estimating Student-t degrees of freedom parameter

The Student-t distribution is commonly used in statistical modeling to handle data with heavier tails than the normal distribution. An essential parameter of the Student-t distribution is the degrees of freedom $\nu$, which controls the tail heaviness. In this note, we present the Bayesian estimation of the degrees of freedom parameter for an N-variate Student-t distribution using the Inverted-Gamma 2 (IG2) scale mixture of normals.

The N-variate Student-t distribution can be represented as a scale mixture of normals:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

where:

- $\mathbf{y}$ is the $N$-dimensional observation vector.
- $\mathbf{\mu}$ is the mean vector.
- $\lambda$ is the latent scale variable.
- $\nu$ is the degrees of freedom parameter.

### Derivation of Full Conditional Posteriors

#### Full Conditional Posterior of $\lambda$

Given the prior distribution:

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

The likelihood of the data given $\lambda$ is:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

The full conditional posterior of $\lambda$ can be derived as follows:

1. **Likelihood of $\mathbf{y}$ given $\mathbf{\mu}$ and $\lambda$**:

   $$
   p(\mathbf{y} \mid \mathbf{\mu}, \lambda) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right)
   $$

2. **Prior for $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) \propto \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

3. **Joint distribution**:

   $$
   p(\mathbf{y}, \lambda \mid \mathbf{\mu}, \nu) = p(\mathbf{y} \mid \mathbf{\mu}, \lambda) p(\lambda \mid \nu)
   $$

4. **Full conditional posterior**:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right) \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

   Combining terms:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\left(\nu + \frac{N}{2} + 1\right)} \exp\left(-\frac{\nu + \frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2}}{\lambda}\right)
   $$

   This is recognized as the kernel of an Inverted-Gamma 2 distribution:

   $$
   \lambda \mid \mathbf{y}, \mathbf{\mu}, \nu \sim \mathcal{IG2}\left(\nu + N, \nu + (\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})\right)
   $$

#### Full Conditional Posterior of $\nu$

To estimate $\nu$, we use the Metropolis-Hastings algorithm due to its non-standard form. The steps for deriving the full conditional posterior of $\nu$ are as follows:

1. **Likelihood of $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) = \frac{\left(\frac{\nu}{2}\right)^{\nu/2}}{\Gamma(\nu/2)} \lambda^{-\left(\nu/2 + 1\right)} \exp\left(-\frac{\nu}{2\lambda}\right)
   $$

2. **Log-likelihood for $\nu$ given $\lambda$**:

   $$
   \log p(\lambda \mid \nu) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

3. **Log-prior for $\nu$ (assuming a non-informative prior)**:

   $$
   \log p(\nu) = \text{constant}
   $$

4. **Full conditional posterior**:

   The full conditional posterior for $\nu$ is proportional to the product of the likelihood and the prior:

   $$
   p(\nu \mid \lambda) \propto p(\lambda \mid \nu) p(\nu)
   $$

   Since $p(\nu)$ is constant, we focus on $p(\lambda \mid \nu)$:

   $$
   \log p(\nu \mid \lambda) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

   This expression does not have a closed form, so we use the Metropolis-Hastings algorithm to sample from this posterior.

### R Function for Gibbs Sampler

Below is the R function implementing the Gibbs sampler for estimating $\nu$ using the IG2-scale mixture of normals representation.

```{r echo=TRUE}
metropolis_hastings_nu <- function(y, mu, n_iter, init_nu, proposal_sd) {
  # Initialize parameter
  nu <- init_nu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  
  # Log-likelihood function
  log_likelihood <- function(nu, y, mu) {
    sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
  }
  
  for (i in 1:n_iter) {
    # Propose new value for nu
    nu_proposal <- nu + rnorm(1, 0, proposal_sd)
    
    if (nu_proposal > 0) {
      # Calculate log acceptance ratio
      log_acceptance_ratio <- log_likelihood(nu_proposal, y, mu) - log_likelihood(nu, y, mu)
      
      # Accept or reject the proposal
      if (log(runif(1)) < log_acceptance_ratio) {
        nu <- nu_proposal
      }
    }
    
    # Store the sample
    nu_samples[i] <- nu
  }
  
  return(nu_samples)
}

gibbs_sampler_t <- function(y, n_iter, init_values) {
  # Initialize parameters
  nu <- init_values$nu
  mu <- init_values$mu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  mu_samples <- numeric(n_iter)
  lambda_samples <- numeric(n_iter)
  
  for (i in 1:n_iter) {
    # Sample lambda
    shape_lambda <- nu + N
    rate_lambda <- nu + sum((y - mu)^2)
    lambda <- 1 / rgamma(1, shape = shape_lambda, rate = rate_lambda)
    
    # Sample mu
    mu <- rnorm(1, mean = mean(y), sd = sqrt(lambda / N))
    
    # Sample nu using Metropolis-Hastings
    log_likelihood <- function(nu, y, mu) {
      sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
    }
    
    proposal_nu <- nu + rnorm(1, 0, 0.1) # proposal distribution: normal random walk
    if (proposal_nu > 0) {
      log_acceptance_ratio <- log_likelihood(proposal_nu, y, mu) - log_likelihood(nu, y, mu)
      if (log(runif(1)) < log_acceptance_ratio

) {
        nu <- proposal_nu
      }
    }
    
    # Store samples
    nu_samples[i] <- nu
    mu_samples[i] <- mu
    lambda_samples[i] <- lambda
  }
  
  return(list(nu = nu_samples, mu = mu_samples, lambda = lambda_samples))
}

# Example usage
set.seed(123)
y <- rnorm(100)
init_values <- list(nu = 5, mu = mean(y))
n_iter <- 1000
result <- gibbs_sampler_t(y, n_iter, init_values)

# Display the results
print(summary(result$nu))
print(summary(result$mu))
print(summary(result$lambda))
```

### Conclusion

This note provided a comprehensive step-by-step algebraic derivation and a sampler for estimating the degrees of freedom parameter $\nu$ for an N-variate Student-t distribution using an IG2-scale mixture of normals approach within a Bayesian framework. By using the Metropolis-Hastings algorithm, we avoid the need to assume a prior distribution for $\nu$, simplifying the estimation process. This approach allows for flexible modeling of heavy-tailed data, which is often encountered in practice.

### References

- Geweke, J. (1993). Bayesian treatment of the independent Student-t linear model. *Journal of Applied Econometrics*, 8(S1), S19-S40.
- Chib, S., & Greenberg, E. (1995). Understanding the Metropolis-Hastings Algorithm. *The American Statistician*, 49(4), 327-335.




## Laplace prior for the trend component

# Model extensions

## Autoregressive cycle component

## Random walk with  time-varying drift parameter

## Student-t error terms

## Conditional heteroskedasticity

# Bayesian forecasting

## Predictive density

## Sampling from the predictive density

## Missing observations


## References {.unnumbered}
