---
title: "Bayesian Unobserved Component Models"
author:
  - name: Tomasz Woźniak
    url: https://github.com/donotdespair
    orcid: 0000-0003-2212-2378
  - name: Ben Gussen
    url: https://github.com/bgussen
    orcid: 0000-0003-4406-4076
  - name: Inhye Kang
    url: https://github.com/inhyekang-au
    orcid: 0009-0007-8544-4952
  - name: Yuqiao Yang
    url: https://github.com/JimmyYang917
  
    execute:
  #echo: false
citation: 
  issued: 2024-05-01
  url: https://donotdespair.github.io/Bayesian-Unobserved-Component-Models/
  doi: 10.26188/25814617
bibliography: references.bib
---

> **Abstract.** We present the basics of Bayesian estimation and inference for unobserved component models on the example of a local-level model. 
> The range of topics includes the conjugate prior analysis using normal-inverted-gamma 2 distribution and its extensions focusing on hierarchical modelling, conditional heteroskedasticity, and Student-t error terms. 
> We scrutinise Bayesian forecasting and sampling from the predictive density.
>
> **Keywords.** Unobserved Component Models, Local-Level Model, State-Space Bayesian Inference, Forecasting, Heteroskedasticity, Hierarchical Modelling, Gibbs Sampler, Simulation Smoother, Precision Sampling

# Unobserved component models

Unobserved Component (UC) models are a popular class of models in macroeconometrics that use the state-space representation for unit-root nonstationary time series. 
The simple formulation of the model equations decomposing the series into a non-stationary and stationary component facilitates economic interpretations and good forecasting performance.

# A simple local-level model

The model is set for a univariate time series whose observation at time
$t$ is denoted by $y_t$. 
It decomposes the variable into a stochastic trend component, $\tau_t$, and a stationary error component, $\epsilon_t$. 
The former follows a Gaussian random walk process with the conditional variance $\sigma_\eta^2$, and the latter is zero-mean normally distributed with the variance $\sigma^2$.
These are expressed as the model equations:
\begin{align}
y_t &= \tau_t + \epsilon_t,\\
\tau_t &= \tau_{t-1} + \eta_t,\\
\epsilon_t &\sim\mathcal{N}\left(0, \sigma^2\right),\\
\eta_t &\sim\mathcal{N}\left(0, \sigma_\eta^2\right),
\end{align}
where the initial condition $\tau_0$ is a parameter of the model.

## Matrix notation for the model

To simplify the notation and the derivations introduce matrix notation for the model. Let $T$ be the available sample size for the variable $y$. 
Define a $T$-vector of zeros, $\mathbf{0}_T$, and of ones, $\boldsymbol\imath_T$, the identity matrix of order $T$, $\mathbf{I}_T$, as well as $T\times1$ vectors:
\begin{align}
\mathbf{y} = \begin{bmatrix} y_1\\ \vdots\\ y_T \end{bmatrix},\quad
\boldsymbol\tau = \begin{bmatrix} \tau_1\\ \vdots\\ \tau_T \end{bmatrix},\quad
\boldsymbol\epsilon = \begin{bmatrix} \epsilon_1\\ \vdots\\ \epsilon_T \end{bmatrix},\quad
\boldsymbol\eta = \begin{bmatrix} \eta_1\\ \vdots\\ \eta_T \end{bmatrix},\qquad
\mathbf{i} = \begin{bmatrix} 1\\0\\ \vdots\\ 0 \end{bmatrix},
\end{align}
and a $T\times T$ matrix $\mathbf{H}$ with the elements:
\begin{align}
\mathbf{H} = \begin{bmatrix}
1 & 0 & \cdots & 0 & 0\\
-1 & 1 & \cdots & 0 & 0\\
0 & -1 & \cdots & 0 & 0\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
0 & 0 & \cdots & 1 & 0\\
0 & 0 & \cdots & -1 & 1
\end{bmatrix}.
\end{align}

Then the model can be written in a concise notation as:
\begin{align}
\mathbf{y} &= \mathbf{\tau} + \boldsymbol\epsilon,\\
\mathbf{H}\boldsymbol\tau &= \mathbf{i} \tau_0 + \boldsymbol\eta,\\
\boldsymbol\epsilon &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma^2\mathbf{I}_T\right),\\
\boldsymbol\eta &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma_\eta^2\mathbf{I}_T\right).
\end{align}

## Likelihood function

The model equations imply the predictive density of the data vector
$\mathbf{y}$. To see this, consider the model equation as a linear
transformation of a normal vector $\boldsymbol\epsilon$. Therefore, the data vector follows a multivariate normal distribution given by:
\begin{align}
\mathbf{y}\mid \boldsymbol\tau, \sigma^2 &\sim\mathcal{N}_T\left(\boldsymbol\tau, \sigma^2\mathbf{I}_T\right).
\end{align}

This distribution determines the shape of the likelihood function that
is defined as the sampling data density: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y})\equiv p\left(\mathbf{y}\mid\boldsymbol\tau, \sigma^2 \right).
\end{align}

The likelihood function that for the sake of the estimation of the parameters, and after plugging in data in place of $\mathbf{y}$, is considered a function of parameters $\boldsymbol\tau$ and $\sigma^2$ is given by: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y}) = 
(2\pi)^{-\frac{T}{2}}\left(\sigma^2\right)^{-\frac{T}{2}}\exp\left\{-\frac{1}{2}\frac{1}{\sigma^2}(\mathbf{y} - \boldsymbol\tau)'(\mathbf{y} - \boldsymbol\tau)\right\}.
\end{align}

## Prior distributions

The state equation for $\boldsymbol\tau$ can be rewritten as follows:
$$
\begin{gather}
\boldsymbol\tau = \mathbf{H}^{-1} \mathbf{i} \tau_0+\mathbf{H}^{-1}\boldsymbol\eta \\
\\ \boldsymbol\eta \sim \mathcal{N}(\mathbf{0}_T, \sigma_\eta^2 \mathbf{I}_T) \\
\mathbf{H}^{-1} \boldsymbol\eta \sim \mathcal{N}(\mathbf{0}_T, \sigma_\eta^2\left(\mathbf{H}^{\prime} \mathbf{H}\right)^{-1})
\end{gather}
$$
Using the state equation for $\boldsymbol\tau$ above, we can derive the prior distribution of $\boldsymbol\tau$ as:
$$
\begin{align}
\boldsymbol\tau | \tau_0, \sigma_\eta^2 &\sim \mathcal{N}_T(\mathbf{H}^{-1} \mathbf{i} \tau_0, \sigma_\eta^2(\mathbf{H}^{\prime} \mathbf{H})^{-1})
\\ &\propto \exp \left\{-\frac{1}{2} \frac{1}{\sigma_\eta^2}\left(\boldsymbol\tau-\mathbf{H}^{-1} \mathbf{i} \tau_0\right)^{\prime}\left(\mathbf{H}^{\prime} \mathbf{H}\right)\left(\boldsymbol\tau-\mathbf{H}^{-1} \mathbf{i} \tau_0\right)\right\}
\end{align}
$$
where the prior distribution of $\tau_0$ is derived as follows:
$$
\begin{align}
\tau_0 &\sim \mathcal{N}(\underline{\tau_0}, \underline{V_{\tau_{0}}}) 
\\ &\propto \exp \left\{-\frac{1}{2}(\tau_0-\underline{\tau_0})^{\prime} \underline{V_{\tau_{0}}}^{-1}(\tau_0-\underline{\tau_0})\right\}
\end{align}
$$
and the prior distribution of $\sigma_\eta^2$ is the following:
$$
\begin{align}
\sigma_\eta^2 &\sim \mathcal{IG}2(\underline{s_\eta}, \underline{v_\eta}) \\
&\propto (\sigma_\eta^2)^{-\frac{\underline{v}+2}{2}} \exp \left\{-\frac{1}{2} \frac{s}{\sigma_\eta^2}\right\}
\end{align}
$$

Given that $\boldsymbol\epsilon$ follows normal distribution,
$$
\begin{align}
\boldsymbol\epsilon &\sim \mathcal{N}(\mathbf{0}_T, \sigma^2 \mathbf{I}_T)
\end{align}
$$
the prior distribution of $\boldsymbol\epsilon$ given $\sigma^2$ is the following:
$$
\begin{align}
\boldsymbol\epsilon | \sigma^2 &\sim \mathcal{N}(\mathbf{0}_T, \sigma^2 \mathbf{I}_T)
\\ &\propto \exp \left\{-\frac{1}{2} \frac{1}{\sigma^2} \boldsymbol\epsilon^{\prime}\boldsymbol\epsilon\right\}
\end{align}
$$

The prior distribution of $\sigma^2$ is the following:
$$
\begin{align}
\sigma^2 &\sim \mathcal{IG}2(\underline{s}, \underline{v}) \\
&\propto (\sigma^2)^{-\frac{\underline{v}+2}{2}} \exp \left\{-\frac{1}{2} \frac{s}{\sigma^2}\right\}
\end{align}
$$

The Joint prior distribution of $\boldsymbol\tau$, $\tau_0$, $\sigma^2_{\eta}$, and $\sigma^2$ is then derived as follows:
$$
\begin{aligned}
p(\boldsymbol\tau,\tau_{0},\sigma_{\eta}^{2},\sigma^{2}) = p(\boldsymbol\tau|\tau_{0},\sigma_{\eta}^{2}) \space p(\tau_{0}) \space p(\sigma_{\eta}^{2}) \space p(\sigma^{2})
\end{aligned}
$$

# Bayesian estimation

The Baye's rule is used to derive full conditional posterior distribution.
$$
\begin{aligned}
p(A|B) & =\frac{p(B|A)p(A)}{p(B)} \\
\\ p(A|B) & \propto p(B|A) \space p(A)
\end{aligned}
$$

Two likelihood functions to derive the full Conditional posterior distribution of $\tau,\tau_{0},\sigma^{2},\sigma_{\eta}^{2}$ is as follows:
$$
\begin{align}
L(\boldsymbol\tau, \sigma^{2}|\mathbf{y}) &\propto \exp\left\{ -\frac{1}{2}\sigma^{-2}\left( \mathbf{y}-\boldsymbol\tau \right)'\left( \mathbf{y}-\boldsymbol\tau \right) \right\}
\end{align}
$$
and
$$
$$

Given the likelihood and priors above, the full Conditional posterior distribution of $\boldsymbol\tau$ is derived as follows:
$$
\begin{align}
P(\boldsymbol\tau|\mathbf{y},\tau_{0},\sigma^{2},\sigma_{\eta}^{2})&\propto L(\boldsymbol\tau, \sigma^{2}|\mathbf{y}) \space P(\boldsymbol\tau|\tau_{0},\sigma_{\eta}^{2})\\
\\&\propto \exp\left\{ -\frac{1}{2}\sigma^{-2}\left( \mathbf{y}-\boldsymbol\tau \right)'\left( \mathbf{y}-\boldsymbol\tau \right) \right\}\\
&\times \exp\left\{ -\frac{1}{2}\sigma_{\eta}^{-2}\left(\boldsymbol\tau-\mathbf{H}^{-1} \mathbf{i}\tau_{0}\right)'\mathbf{H}'\mathbf{H}\left( \boldsymbol\tau-\mathbf{H}^{-1} \mathbf{i}\tau_{0} \right) \right\}
\end{align}
$$
where
$$
\begin{align}
P(\boldsymbol\tau &|\mathbf{y},\tau_{0},\sigma^{2},\sigma_{\eta}^{2}) = \mathcal{N}(\overline{\boldsymbol\tau},\overline{V_{\boldsymbol\tau}}) \\ 
\\ \overline{V_{\boldsymbol\tau}} &= \left[ \sigma^{-2}\mathbf{I}_T +\sigma_{\eta}^{-2}\mathbf{H}'\mathbf{H}\right]^{-1}\\
\overline{\boldsymbol\tau} &= \overline{V_{\tau}}\left[\sigma^{-2}\mathbf{y} + \sigma_{\eta}^{-2}\mathbf{H}'\mathbf{i}\tau_{0} \right]
\end{align}
$$

The full conditional posterior distribution of $\tau_{0}$ is the following:
$$
\begin{align}
P(\tau_{0}|\mathbf{y},\boldsymbol\tau,\sigma_{\eta}^{2}) &\propto L(\tau_{0}|\mathbf{y},\boldsymbol\tau,\sigma_{\eta}^{2}) \space P(\tau_{0}) \\
\\ &\propto \exp\left\{ -\frac{1}{2}\sigma_{\eta}^{-2}\left(\mathbf{H}\boldsymbol\tau -\mathbf{i}\tau_{0}\right)'\left(\mathbf{H}\boldsymbol\tau-\mathbf{i}\tau_{0}\right)\right\}\\
&\times \exp\left\{ -\frac{1}{2}\left(\tau_{0}-  \underline{\tau_0}\right)'\underline{V_{\tau_{0}}}^{-1}\left(\tau_{0}-  \underline{\tau_0}\right) \right\}
\end{align}
$$
where
$$
\begin{align}
P(\tau_{0} &|\mathbf{y},\boldsymbol\tau,\sigma_{\eta}^{2}) = N(\overline{\tau_{0}},\overline{V_{\tau_{0}}}) \\
\\ \overline{V_{\tau_{0}}}&=\left[ \sigma_{\eta}^{-2} +\underline{V_{\tau_{0}}}^{-1}\right]^{-1} \\
\overline{\tau_{0}}&=\overline{V_{\tau_{0}}}\left[ \sigma_{\eta}^{-2} \mathbf{i}'\mathbf{H}\boldsymbol\tau +\underline{V_{\tau_{0}}}^{-1}\underline{\tau_{0}}\right]
\end{align}
$$
The full conditional posterior distribution of $\sigma_{\eta}^{2}$ is derived as:
$$
\begin{align}
P(\sigma_{\eta}^{2}|\mathbf{y},\boldsymbol\tau,\tau_{0}) &\propto L(\sigma_{\eta}^{2}|\mathbf{y},\boldsymbol\tau,\tau_{0}) \space P(\sigma_{\eta}^{2}) \\
\\ &\propto \left( \sigma_{\eta}^{2} \right)^{-\frac{T}{2}} \exp\left\{ -\frac{1}{2}\sigma_{\eta}^{-2}\left(\mathbf{H}\boldsymbol\tau -\mathbf{i}\tau_{0}\right)'\left(\mathbf{H}\boldsymbol\tau -\mathbf{i}\tau_{0}\right)\right\}
\\ &\times \left( \sigma_{\eta}^{2} \right)^{-\frac{\underline{\upsilon}+2}{2}} \exp\left\{ -\frac{1}{2}\frac{\underline{s}}{\sigma_{\eta}^{2}}\right\}
\end{align}
$$
where
$$
\begin{align}
P(\sigma_{\eta}^{2}&|\mathbf{y},\boldsymbol\tau,\tau_{0}) = \mathcal{IG}2(\overline{s_{\eta}},\overline{\upsilon_{\eta}}) \\
\\ \overline{s_{\eta}}&=\underline{s} + \left(\mathbf{H}\boldsymbol\tau -\mathbf{i}\tau_{0}\right)'\left(\mathbf{H}\boldsymbol\tau -\mathbf{i}\tau_{0}\right) \\
\overline{\upsilon_{\eta}}&= \underline{\upsilon}+T
\end{align}
$$
Finally, the full conditional posterior distribution of $\sigma^{2}$ is the following:
$$
\begin{align}
P(\sigma^{2}|\mathbf{y},\boldsymbol\tau)&=L(\boldsymbol\tau,\sigma^{2}|\mathbf{y}) \space P(\sigma^{2}) \\
\\ &\propto \left( \sigma^{2} \right)^{-\frac{T}{2}} \exp\left\{ -\frac{1}{2}\sigma^{-2}\left(\mathbf{y}-\boldsymbol\tau\right)'\left(\mathbf{y}-\boldsymbol\tau\right)\right\}
\\ &\times \left( \sigma^{2} \right)^{-\frac{\underline{\upsilon}+2}{2}} \exp\left\{ -\frac{1}{2}\frac{\underline{s}}{\sigma^{2}}\right\}
\end{align}
$$
where
$$
\begin{align}
P(\sigma^{2}&|\mathbf{y},\boldsymbol\tau) = \mathcal{IG}2(\overline{s},\overline{\upsilon}) \\
\\ \overline{s}&=\underline{s} + \left(\mathbf{y}-\boldsymbol\tau\right)'\left(\mathbf{y}-\boldsymbol\tau\right) \\
\overline{\upsilon}&= \underline{\upsilon}+T
\end{align}
$$

## Gibbs sampler

A **Gibbs Sampler** can be applied using the following steps:

At each iteration $s$ where $s$ goes from 1 to $S$,

|    Step 1. Draw $\tau_0^{(s)}$ from the $N({\tau_{0}}, {V_{\tau_{0}}})$ distribution and collect $\tau_0$.
|    Step 2. Draw ${\sigma^2_{\eta}}^{(s)}$ from the $\mathcal{IG}2({S_\eta}, {v_\eta})$ distribution and collect $\sigma_\eta^2$.
|    Step 3. Draw ${\sigma^2}^{(s)}$ from the $\mathcal{IG}2(S, v)$ distribution and collect $\sigma^2$.
|    Step 4. Draw $\boldsymbol\tau^{(s)}$ from the $\mathcal{N}({\boldsymbol\tau},{V_{\tau}})$ distribution and collect $\boldsymbol\tau$.

```{r Gibbs Sampler}
#| echo: true
# Gibbs sampler for a simple UC model using simulation smoother
############################################################

UC.Gibbs.sampler    = function(S, starting.values, priors){
  # Initialise the data 
  aux     = starting.values
  T       = nrow(aux$Y)
  i_matrix <- diag(T)
  i <- matrix(0, T, 1)  
  i[1, 1] <- 1 
  # Posteriors list
  posteriors    = list(
    tau     = matrix(NA,T,S),
    tau_0   = matrix(NA,1,S),
    sigma   = matrix(NA,2,S),
    non.stationary.iterations = rep(NA,S)
  )
  
  for (s in 1:S){
    
    # Sampling tau_0
    ###########################
    tau_0.v.inv    = diag(1/diag(priors$tau_0.v))
    V.tau_0.bar    = solve((1/aux$sigma[1])*i_matrix + tau_0.v.inv )
    tau_0.bar      = V.tau_0.bar %*% ( (1/aux$sigma[1])*i_matrix%*%priors$H%*%aux$tau%*%i + tau_0.v.inv%*%priors$tau_0 )
    tau_0.draw     = rmvnorm(1,as.vector(tau_0.bar),V.tau_0.bar)
    aux$tau_0      = as.vector(tau_0.draw)
    
    # Sampling sigma
    ###########################
    # sigma of tau (sigma_eta)
    sigma.eta.s   = as.numeric(priors$sigma.s + crossprod(priors$H%*%aux$tau - i%*%priors$tau_0))
    sigma.eta.nu  = priors$sigma.nu + T
    sigma.eta.draw = sigma.eta.s/rchisq(1,sigma.eta.nu)
    # sigma of errors (sigma)
    sigma.e.s     = as.numeric(priors$sigma.s + crossprod(aux$Y - aux$tau))
    sigma.e.nu    = priors$sigma.nu + T
    sigma.e.draw  = sigma.e.s/rchisq(1,sigma.e.nu)
    aux$sigma     = c(sigma.eta.draw,sigma.e.draw)
    
    # Sampling tau
    ###########################
    V.tau.inv     = (1/aux$sigma[2])*i_matrix + (1/aux$sigma[1])*crossprod(priors$H)
    V.tau.inv     = 0.5*(V.tau.inv + t(V.tau.inv))
    b.tau         = (1/aux$sigma[2])*i_matrix%*%aux$Y + (1/aux$sigma[1])%*%i_matrix%*%t(priors$H)%*%i%*%priors$tau_0
    precision.L   = t(bandchol(V.tau.inv))
    epsilon       = rnorm(T)
    b.tau.tmp     = forwardsolve(precision.L, b.tau)
    tau.draw      = backsolve(t(precision.L), b.tau.tmp + epsilon)
    aux$tau       = tau.draw
    
    posteriors$tau[,s]     = aux$tau
    posteriors$tau_0[,s]   = aux$tau_0
    posteriors$sigma[,s]   = aux$sigma
    # posteriors$non.stationary.iterations[s] = ns.i
    if (s%%1000==0){cat(" ",s)}
  }
  
  output      = list(
    posterior = posteriors,
    last.draw = aux
  )
  return(output)
}
```


## Simulation smoother and precision sampler

## Analytical solution for a joint posterior

# Hierarchical modeling

## Estimating gamma error term variance prior scale

## Estimating inverted-gamma 2 error term variance prior scale

## Estimating the initial condition prior scale

## Student-t prior for the trend component

## Estimating Student-t degrees of freedom parameter

The Student-t distribution is commonly used in statistical modeling to handle data with heavier tails than the normal distribution. An essential parameter of the Student-t distribution is the degrees of freedom $\nu$, which controls the tail heaviness. In this note, we present the Bayesian estimation of the degrees of freedom parameter for an N-variate Student-t distribution using the Inverted-Gamma 2 (IG2) scale mixture of normals.

The N-variate Student-t distribution can be represented as a scale mixture of normals:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

where:

- $\mathbf{y}$ is the $N$-dimensional observation vector.
- $\mathbf{\mu}$ is the mean vector.
- $\lambda$ is the latent scale variable.
- $\nu$ is the degrees of freedom parameter.

### Derivation of Full Conditional Posteriors

#### Full Conditional Posterior of $\lambda$

Given the prior distribution:

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

The likelihood of the data given $\lambda$ is:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

The full conditional posterior of $\lambda$ can be derived as follows:

1. **Likelihood of $\mathbf{y}$ given $\mathbf{\mu}$ and $\lambda$**:

   $$
   p(\mathbf{y} \mid \mathbf{\mu}, \lambda) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right)
   $$

2. **Prior for $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) \propto \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

3. **Joint distribution**:

   $$
   p(\mathbf{y}, \lambda \mid \mathbf{\mu}, \nu) = p(\mathbf{y} \mid \mathbf{\mu}, \lambda) p(\lambda \mid \nu)
   $$

4. **Full conditional posterior**:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right) \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

   Combining terms:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\left(\nu + \frac{N}{2} + 1\right)} \exp\left(-\frac{\nu + \frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2}}{\lambda}\right)
   $$

   This is recognized as the kernel of an Inverted-Gamma 2 distribution:

   $$
   \lambda \mid \mathbf{y}, \mathbf{\mu}, \nu \sim \mathcal{IG2}\left(\nu + N, \nu + (\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})\right)
   $$

#### Full Conditional Posterior of $\nu$

To estimate $\nu$, we use the Metropolis-Hastings algorithm due to its non-standard form. The steps for deriving the full conditional posterior of $\nu$ are as follows:

1. **Likelihood of $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) = \frac{\left(\frac{\nu}{2}\right)^{\nu/2}}{\Gamma(\nu/2)} \lambda^{-\left(\nu/2 + 1\right)} \exp\left(-\frac{\nu}{2\lambda}\right)
   $$

2. **Log-likelihood for $\nu$ given $\lambda$**:

   $$
   \log p(\lambda \mid \nu) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

3. **Log-prior for $\nu$ (assuming a non-informative prior)**:

   $$
   \log p(\nu) = \text{constant}
   $$

4. **Full conditional posterior**:

   The full conditional posterior for $\nu$ is proportional to the product of the likelihood and the prior:

   $$
   p(\nu \mid \lambda) \propto p(\lambda \mid \nu) p(\nu)
   $$

   Since $p(\nu)$ is constant, we focus on $p(\lambda \mid \nu)$:

   $$
   \log p(\nu \mid \lambda) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

   This expression does not have a closed form, so we use the Metropolis-Hastings algorithm to sample from this posterior.

### R Function for Gibbs Sampler

Below is the R function implementing the Gibbs sampler for estimating $\nu$ using the IG2-scale mixture of normals representation.

```{r echo=TRUE}
metropolis_hastings_nu <- function(y, mu, n_iter, init_nu, proposal_sd) {
  # Initialize parameter
  nu <- init_nu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  
  # Log-likelihood function
  log_likelihood <- function(nu, y, mu) {
    sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
  }
  
  for (i in 1:n_iter) {
    # Propose new value for nu
    nu_proposal <- nu + rnorm(1, 0, proposal_sd)
    
    if (nu_proposal > 0) {
      # Calculate log acceptance ratio
      log_acceptance_ratio <- log_likelihood(nu_proposal, y, mu) - log_likelihood(nu, y, mu)
      
      # Accept or reject the proposal
      if (log(runif(1)) < log_acceptance_ratio) {
        nu <- nu_proposal
      }
    }
    
    # Store the sample
    nu_samples[i] <- nu
  }
  
  return(nu_samples)
}

gibbs_sampler_t <- function(y, n_iter, init_values) {
  # Initialize parameters
  nu <- init_values$nu
  mu <- init_values$mu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  mu_samples <- numeric(n_iter)
  lambda_samples <- numeric(n_iter)
  
  for (i in 1:n_iter) {
    # Sample lambda
    shape_lambda <- nu + N
    rate_lambda <- nu + sum((y - mu)^2)
    lambda <- 1 / rgamma(1, shape = shape_lambda, rate = rate_lambda)
    
    # Sample mu
    mu <- rnorm(1, mean = mean(y), sd = sqrt(lambda / N))
    
    # Sample nu using Metropolis-Hastings
    log_likelihood <- function(nu, y, mu) {
      sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
    }
    
    proposal_nu <- nu + rnorm(1, 0, 0.1) # proposal distribution: normal random walk
    if (proposal_nu > 0) {
      log_acceptance_ratio <- log_likelihood(proposal_nu, y, mu) - log_likelihood(nu, y, mu)
      if (log(runif(1)) < log_acceptance_ratio

) {
        nu <- proposal_nu
      }
    }
    
    # Store samples
    nu_samples[i] <- nu
    mu_samples[i] <- mu
    lambda_samples[i] <- lambda
  }
  
  return(list(nu = nu_samples, mu = mu_samples, lambda = lambda_samples))
}

# Example usage
set.seed(123)
y <- rnorm(100)
init_values <- list(nu = 5, mu = mean(y))
n_iter <- 1000
result <- gibbs_sampler_t(y, n_iter, init_values)

# Display the results
print(summary(result$nu))
print(summary(result$mu))
print(summary(result$lambda))
```

### Conclusion

This note provided a comprehensive step-by-step algebraic derivation and a sampler for estimating the degrees of freedom parameter $\nu$ for an N-variate Student-t distribution using an IG2-scale mixture of normals approach within a Bayesian framework. By using the Metropolis-Hastings algorithm, we avoid the need to assume a prior distribution for $\nu$, simplifying the estimation process. This approach allows for flexible modeling of heavy-tailed data, which is often encountered in practice.

### References

- Geweke, J. (1993). Bayesian treatment of the independent Student-t linear model. *Journal of Applied Econometrics*, 8(S1), S19-S40.
- Chib, S., & Greenberg, E. (1995). Understanding the Metropolis-Hastings Algorithm. *The American Statistician*, 49(4), 327-335.




## Laplace prior for the trend component

# Model extensions

## Autoregressive cycle component

## Random walk with  time-varying drift parameter

## Student-t error terms

## Conditional heteroskedasticity

# Bayesian forecasting

## Predictive density

## Sampling from the predictive density

## Missing observations


## References {.unnumbered}
