---
title: "Bayesian Unobserved Component Models"
author:
  - name: Tomasz Woźniak
    url: https://github.com/donotdespair
    orcid: 0000-0003-2212-2378
  - name: Ben Gussen
    url: https://github.com/bgussen
    orcid: 0000-0003-4406-4076
  - name: Rui Liu
    url: https://github.com/rruiliu1005
    orcid: 0009-0008-9348-8581
  
    execute:
  #echo: false
citation: 
  issued: 2024-05-01
  url: https://donotdespair.github.io/Bayesian-Unobserved-Component-Models/
  doi: 10.26188/25814617
bibliography: references.bib
---

> **Abstract.** We present the basics of Bayesian estimation and inference for unobserved component models on the example of a local-level model. 
> The range of topics includes the conjugate prior analysis using normal-inverted-gamma 2 distribution and its extensions focusing on hierarchical modelling, conditional heteroskedasticity, and Student-t error terms. 
> We scrutinise Bayesian forecasting and sampling from the predictive density.
>
> **Keywords.** Unobserved Component Models, Local-Level Model, State-Space Bayesian Inference, Forecasting, Heteroskedasticity, Hierarchical Modelling, Gibbs Sampler, Simulation Smoother, Precision Sampling

# Unobserved component models

Unobserved Component (UC) models are a popular class of models in macroeconometrics that use the state-space representation for unit-root nonstationary time series. 
The simple formulation of the model equations decomposing the series into a non-stationary and stationary component facilitates economic interpretations and good forecasting performance.

# A simple local-level model

The model is set for a univariate time series whose observation at time
$t$ is denoted by $y_t$. 
It decomposes the variable into a stochastic trend component, $\tau_t$, and a stationary error component, $\epsilon_t$. 
The former follows a Gaussian random walk process with the conditional variance $\sigma_\eta^2$, and the latter is zero-mean normally distributed with the variance $\sigma^2$.
These are expressed as the model equations:
\begin{align}
y_t &= \tau_t + \epsilon_t,\\
\tau_t &= \tau_{t-1} + \eta_t,\\
\epsilon_t &\sim\mathcal{N}\left(0, \sigma^2\right),\\
\eta_t &\sim\mathcal{N}\left(0, \sigma_\eta^2\right),
\end{align}
where the initial condition $\tau_0$ is a parameter of the model.

## Matrix notation for the model

To simplify the notation and the derivations introduce matrix notation for the model. Let $T$ be the available sample size for the variable $y$. 
Define a $T$-vector of zeros, $\mathbf{0}_T$, and of ones, $\boldsymbol\imath_T$, the identity matrix of order $T$, $\mathbf{I}_T$, as well as $T\times1$ vectors:
\begin{align}
\mathbf{y} = \begin{bmatrix} y_1\\ \vdots\\ y_T \end{bmatrix},\quad
\boldsymbol\tau = \begin{bmatrix} \tau_1\\ \vdots\\ \tau_T \end{bmatrix},\quad
\boldsymbol\epsilon = \begin{bmatrix} \epsilon_1\\ \vdots\\ \epsilon_T \end{bmatrix},\quad
\boldsymbol\eta = \begin{bmatrix} \eta_1\\ \vdots\\ \eta_T \end{bmatrix},\qquad
\mathbf{i} = \begin{bmatrix} 1\\0\\ \vdots\\ 0 \end{bmatrix},
\end{align}
and a $T\times T$ matrix $\mathbf{H}$ with the elements:
\begin{align}
\mathbf{H} = \begin{bmatrix}
1 & 0 & \cdots & 0 & 0\\
-1 & 1 & \cdots & 0 & 0\\
0 & -1 & \cdots & 0 & 0\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
0 & 0 & \cdots & 1 & 0\\
0 & 0 & \cdots & -1 & 1
\end{bmatrix}.
\end{align}

Then the model can be written in a concise notation as:
\begin{align}
\mathbf{y} &= \mathbf{\tau} + \boldsymbol\epsilon,\\
\mathbf{H}\boldsymbol\tau &= \mathbf{i} \tau_0 + \boldsymbol\eta,\\
\boldsymbol\epsilon &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma^2\mathbf{I}_T\right),\\
\boldsymbol\eta &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma_\eta^2\mathbf{I}_T\right).
\end{align}

## Likelihood function

The model equations imply the predictive density of the data vector
$\mathbf{y}$. To see this, consider the model equation as a linear
transformation of a normal vector $\boldsymbol\epsilon$. Therefore, the data vector follows a multivariate normal distribution given by:
\begin{align}
\mathbf{y}\mid \boldsymbol\tau, \sigma^2 &\sim\mathcal{N}_T\left(\boldsymbol\tau, \sigma^2\mathbf{I}_T\right).
\end{align}

This distribution determines the shape of the likelihood function that
is defined as the sampling data density: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y})\equiv p\left(\mathbf{y}\mid\boldsymbol\tau, \sigma^2 \right).
\end{align}

The likelihood function that for the sake of the estimation of the parameters, and after plugging in data in place of $\mathbf{y}$, is considered a function of parameters $\boldsymbol\tau$ and $\sigma^2$ is given by: 
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y}) = 
(2\pi)^{-\frac{T}{2}}\left(\sigma^2\right)^{-\frac{T}{2}}\exp\left\{-\frac{1}{2}\frac{1}{\sigma^2}(\mathbf{y} - \boldsymbol\tau)'(\mathbf{y} - \boldsymbol\tau)\right\}.
\end{align}

## Prior distributions

# Bayesian estimation

## Gibbs sampler

## Simulation smoother and precision sampler

## Analytical solution for a joint posterior
### 1. Model Setup 
In this section, we present an analytical solution for a joint posterior for \tau and \sigma for a variation to the standard Bayesian unobserbed component model. Specifically, we assume in this case $\tau_0 = 0$ and $\sigma_{\eta}^2 = c\sigma^2$ is constant and c is constant signal-to-noise ratio. We have the following setup: 

$$\underset{T \times 1}{y} = \begin{pmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{pmatrix} \qquad 
\underset{T \times 1}{\tau} = \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\\tau_T \end{pmatrix} \qquad 
\underset{T \times 1}{\epsilon_{1:T}} = \begin{pmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{pmatrix} \qquad 
\underset{T \times 1}{\eta} = \begin{pmatrix} \eta_1 \\ \eta_2 \\ \vdots \\ \eta_T \end{pmatrix}\qquad 
\underset{T \times 1}{e} = \begin{pmatrix} e_1 \\ e_2 \\ \vdots \\ e_T \end{pmatrix} $$
$$\underset{T \times 1}{I_T} = \begin{pmatrix}1 \\ 1 \\ \vdots \\ 1\end{pmatrix} \qquad \underset{T \times 1}{e_{1.T}} = \begin{pmatrix} 1 \\ 0 \\ \vdots \\ 0 \end{pmatrix} \qquad 
\underset{2 \times 1}{\beta} = \begin{pmatrix}\mu \\ 0 \end{pmatrix} \qquad 
\underset{p \times 1}{\alpha} = \begin{pmatrix} \alpha_1 \\ \alpha_2 \\ \vdots \\ \alpha_p \end{pmatrix} \qquad \color{purple}{\mathbf{\underset{T \times 1}{x_{\tau}} = I_T}}$$

$$\underset{T \times p}{X_{\epsilon}} = \begin{bmatrix}\epsilon_{0:T-1} & \epsilon_{-1:T-2} & \cdots & \epsilon_{-p+1:T-p} \end{bmatrix} = \begin{pmatrix} \epsilon_0 & \epsilon_1 & \cdots & \epsilon_{-p+1} \\ \epsilon_1 & \epsilon_2 & \cdots & \epsilon_{-p+2} \\ \vdots & \vdots & \cdots & \vdots \\ \epsilon_{T-1} & \epsilon_{T-2} & \cdots & \epsilon_{T-p} \end{pmatrix}$$ 

$$\underset{T \times T}{H} = 
\begin{pmatrix} 1 & 0 & 0 & \cdots & 0 & 0 \\ 
-1 & 1 & 0 & \cdots & 0 & 0 \\ 
0 & -1 & 1 & \cdots & 0 & 0 \\ 
\vdots & \vdots & \ddots & \ddots & \vdots & \vdots \\ 
0&0&\vdots & \vdots & 1 & 0  \\ 
0 & 0 & 0 &0& -1 & 1 \end{pmatrix} \qquad 
\underset{T \times T}{H_{\alpha}} = 
\begin{pmatrix} 1 & 0 & 0 & 0 & \cdots&\cdots& \cdots&0 & 0 \\ 
-\alpha_1 & 1 &0 & 0 & \cdots & \cdots&\cdots&0 & 0 \\ 
-\alpha_2 & -\alpha_1 & 1 & 0 & \cdots & \cdots&\cdots& 0 & 0 \\ 
- \alpha_3 & -\alpha_2 & -\alpha_1 & 1 & \ddots &  \cdots& \cdots&0 & 0 \\ 
\ddots & \ddots & \ddots & \ddots&\ddots& \ddots & \ddots & \ddots & \ddots \\ 
-\alpha_p & -\alpha_{p-1} & \ddots & \cdots &\cdots& \cdots& \cdots & \cdots & 0 \\ 
0 & -\alpha_p &  \ddots & \ddots &\cdots&\cdots& \ddots & \ddots & 0 \\ 
\vdots & \vdots & \ddots & \ddots & \ddots&\ddots&\ddots & \ddots & \vdots \\ 
0 & 0 & \cdots  & -\alpha_p & \cdots & -\alpha_3& - \alpha_2& -\alpha_1 & 1\end{pmatrix}$$ 

In this setting, we have out state equations as: 

$$y = \tau + \epsilon \qquad \qquad\qquad\qquad \quad\qquad \qquad \qquad \qquad \text{measurement equation}$$

$$\tau = H^{-1}X_{\tau}\beta + H^{-1} \eta  = \color{purple}{\mathbf{H^{-1}\mu I_T + H^{-1}\eta}}  \quad \color{black}{  \qquad \qquad\text{state equation 1}}$$

$$\epsilon = X_{\epsilon} \alpha + e = H^{-1}_{\alpha}e  \qquad\quad \qquad \quad \qquad \qquad \qquad \qquad\text{state equation 2} $$

The state equations in matrix notation are derived as follows: 

(1) In an unobserved component model, the measurement equation is defined as $y_t = \tau_t + \epsilon_t$, in matrix notation, we have: 

$$y = \begin{pmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{pmatrix} = \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\ \tau_T \end{pmatrix} + \begin{pmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{pmatrix} = \tau + \epsilon$$
(2) We have for each t = 1,...,T, $\tau_t = \mu + \tau_{t-1} + \eta_t$, and $\tau_0 = 0$, so in matrix notation: 

$$\tau = \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\ \tau_T \end{pmatrix} = \begin{pmatrix} \mu + \eta_1 \\ 
\mu + \tau_1+ \eta_2 \\
\mu + \tau_2+ \eta_3 \\ 
\vdots \\ 
\mu + \tau_{T-1} + \eta_T  \end{pmatrix} = \mu I_T + \begin{pmatrix} 0 \\ \tau_1 \\ \vdots \\ \tau_{T-1} \end{pmatrix} + \eta$$

Moving the lagged $\tau$ terms on the right hand side to the left:

$$\begin{pmatrix} 
1 & 0 & \cdots & 0 \\
-1 & 1 & \cdots & 0 \\ 
\vdots & \ddots & \ddots & \vdots \\ 
0 & 0 & -1 & 1
\end{pmatrix} \begin{pmatrix} \tau_1 \\ \tau_2 \\ \vdots \\ \tau_T \end{pmatrix}  = \begin{pmatrix} \tau_1 \\ \tau_2-\tau_1 \\ \vdots \\ \tau_T - \tau_{T-1} \end{pmatrix} = \mu I_T + \eta$$

Hence, we have: 
$$H \tau = \mu I_T + \eta = x_{\tau}\beta + \eta \implies \tau = H^{-1}X_\tau\beta + H^{-1}\eta$$

(3) the unit room stationary component is defined as: $\epsilon_t = \alpha_1 \epsilon_{t-1} + \alpha_2 \epsilon_{t-2} + \cdots + \alpha_p \epsilon_{t-p} + e+t$ In matrix, notation: 

$$\begin{pmatrix} 
\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{pmatrix} = \begin{pmatrix} \alpha_1 \epsilon_0 + \alpha_2 \epsilon_{-1} + \cdots+\alpha_p \epsilon_{-p+1} +e_1 \\ 
\alpha_1 \epsilon_1+ \alpha_2 \epsilon_{0} + \cdots+\alpha_p \epsilon_{-p+2} +e_2 \\ 
\vdots  \\
\alpha_1 \epsilon_T + \alpha_2 \epsilon_{T-1} + \cdots+\alpha_p \epsilon_{T-p} +e_T 
\end{pmatrix} \implies H_{\alpha} \epsilon = e \implies \epsilon = H_{\alpha}^{-1}e$$
since we assume $\begin{pmatrix}\epsilon_0, \epsilon_1, \cdots, \epsilon_T\end{pmatrix}^T = 0$
Similarly, we have 

$$\begin{pmatrix} 
\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{pmatrix} = 
\begin{pmatrix} \epsilon_0 & \epsilon_{-1} & \cdots & \epsilon_{-p+1} \\
\epsilon_1 & \epsilon_{0} & \cdots & \epsilon_{-p+2} \\
\vdots & \vdots & \vdots & \vdots \\ 
\epsilon_{T-1} & \epsilon_{T-2} & \cdots & \epsilon_{-p+T} \\
\end{pmatrix} \begin{pmatrix} \alpha_1 \\ \alpha_2 \\ \vdots \\ \alpha_p \end{pmatrix} + \begin{pmatrix}  e_1 \\ e_2 \\ \vdots \\ e_T \end{pmatrix} = X_{\epsilon} \alpha + e$$ 

We have $\mathbf{\color{purple}{\eta \sim \mathcal{N}(0_T, c\sigma^2I_T)}}$ and $e \sim \mathcal{N}(0_T, \sigma^2_eI_T)$. 

The parameters to be estimated are: $\mathbf{\color{purple}{\sigma^2_e, \tau, \epsilon, \beta = \mu}}$

### 2.Prior distributions 

(1) **Prior distribution for** $\tau$ 

From the state equation, we have: 
$$\tau = H^{-1}\mu I_T + H^{-1}\eta$$
we know that $$\mathbf{\color{purple}{\eta \sim \mathcal{N}(0_T, c \sigma^2 I_T) \implies H^{-1}\eta \sim \mathcal{N}(0_T, c\sigma^2(H^TH)^{-1})}}$$
since $Var(H^{-1}\eta) = H^{-1}Var(\eta)(H^{-1})^T =c\sigma^2(H^TH)^{-1}$
Then, the prior distribution of $\tau|\beta, c\sigma^2$ is 
$$\tau|\beta, c\sigma^2 \sim \mathcal{N}(H^{-1}\mu I_T, c\sigma^2(H^TH)^{-1}) \\ \propto exp\left(-\frac{1}{2}\frac{(\tau - H^{-1}\mu I_T)^TH^TH(\tau - H^{-1}\mu I_T)}{c\sigma^2}\right)$$
(2) **Prior distribution for** $\epsilon$
From the state equation, we have: 

$$\epsilon = H^{-1}_{\alpha}e$$ 
since $e \sim \mathcal{N}(0_T, \sigma^2_eI_T)$, we have: 

$$\epsilon | \sigma_e, \alpha \sim \mathcal{N}(0_T,\sigma_e^2(H_\alpha^TH_\alpha)^{-1}) \\ \propto exp\left(\frac{1}{2\sigma^2_e}\epsilon^T(H_\alpha^TH_\alpha)\epsilon\right)$$
(3) **Prior assumptions for other parameters**

$$\alpha \sim \mathcal{N}({\alpha_{prior}, V_{prior}})\mathbb{1}_{\alpha \in A} \propto exp\left({-\frac{1}{2}(\alpha - \alpha_{prior})^T}V_{\alpha,prior}^{-1}(\alpha - \alpha_{prior})\right)\mathbb{1}_{\alpha \in A}$$
$$\beta = \mu\sim \mathcal{N}(\beta_{prior}, V_{\beta, prior} \propto exp\left({-\frac{1}{2}(\beta - \beta_{prior})^T}V_{\beta,prior}^{-1}(\beta - \beta_{prior})\right)$$
$$\sigma^2_e \sim \mathcal{IG2}(s_{prior}, \nu_{prior}) \propto (\sigma^2_e)^\frac{-\nu+2}{2}exp\left(-\frac{s_{prior}}{2\sigma^2_e}\right)$$
and 

$$p(\tau, \epsilon, \alpha, \beta, \sigma^2_e) = p(\tau|\beta)p(\beta)p(\epsilon|\alpha,\sigma^2_e)p(\alpha)p(\sigma^2_e)$$

### 3. Posterior distributions

1. $\mathbf{p(\tau|y, \alpha, \beta, \sigma^2_e)}$
We have, from the measurement equation: 
$$y = \tau + \epsilon$$
since $\epsilon \sim \mathcal{N}(0_T, \sigma^2_e(H^T_\alpha H_\alpha)^{-1})$
$$y|\tau, \alpha, \sigma^2_e \sim \mathcal{N}(\tau, \sigma^2_e(H^T_\alpha H_\alpha)^{-1})\propto exp\left(-\frac{1}{2\sigma^2_e}(y-\tau)^TH_\alpha^TH_\alpha(y-\tau)\right)$$
and the prior of $\tau$ is: 

$$\tau|\beta, c\sigma^2 \sim \mathcal{N}(H^{-1}\mu I_T, c\sigma^2(H^TH)^{-1}) \\ \propto exp\left(-\frac{1}{2}\frac{(\tau - H^{-1}\mu I_T)^TH^TH(\tau - H^{-1}\mu I_T)}{c\sigma^2}\right)$$

The posterior of $\tau$ is: 

$$p(\tau |\alpha,  \beta, \sigma^2_e, y) = \frac{p(\tau,\alpha, \beta, \sigma^2_e, y)}{p(\alpha, \beta, \sigma^2_e, y)} \propto p(\tau,\alpha, \beta, \sigma^2_e, y)=p(y|\tau, \alpha, , \beta, \sigma^2_e)p(\tau, \alpha, \beta, \sigma^2_e)$$

$$= p(y|\tau, \alpha, \beta, \sigma^2_e)p(\tau|\alpha, \beta, \sigma^2_e)p(\alpha, \beta, \sigma^2_e) \propto  p(y|\tau, \alpha, \beta, \sigma^2_e)p(\tau|\alpha, \beta, \sigma^2_e)$$

$$\propto exp\left(-\frac{1}{2\sigma^2_e}(y-\tau)^TH_\alpha^TH_\alpha(y-\tau)\right)exp\left(-\frac{1}{2}\frac{(\tau - H^{-1}\mu I_T)^TH^TH(\tau - H^{-1}\mu I_T)}{c\sigma^2}\right)$$

$$= exp\left(-\frac{1}{2}\left[\sigma^{-2}_ey^T H_\alpha^TH_\alpha y - 2\sigma_e^{-2}\tau^TH_\alpha^TH_\alpha y +  \sigma_e^{-2}\tau^T H_\alpha^TH_\alpha \tau 
+ c^{-1}\sigma^{-2}\tau^T H^TH \tau -2 c^{-1}\sigma^{-2} \tau^T H^T H H^{-1} \mu I_T + \mu I_T^T (H^{-1})^TH^THH^{-1}\mu I_T
\right]\right)$$

$$\propto exp\left(-\frac{1}{2}\left[ 
\tau^T(\sigma_e^{-2}H_\alpha^TH_\alpha + c^{-1}\sigma^{-2}H^TH)\tau - 2\tau^T(\sigma_e^{-2}H_\alpha^TH_\alpha y + c^{-1}\sigma^{-2} H^T \mu I_T)
\right]\right)$$

Hence, $\tau|\alpha, \beta, \sigma^2_e, y \sim \mathcal{N}(\bar{\tau}, \bar{V_\tau})$ where, 

$$\bar{V_\tau} = [(\sigma_e^{-2}H_\alpha^TH_\alpha+\color{purple}{\mathbf{c^{-1}\sigma^{-2}H^T H)}}]^{-1}$$
$$\bar{\tau} =  \bar{V_\tau}(\sigma_e^{-2}H_\alpha^TH_\alpha y + \color{purple}{\mathbf{c^{-1}\sigma^{-2}H^T \mu I_T}})$$
2. $\mathbf{p(\epsilon|y, \alpha, \beta, \sigma^2_e)}$
We have, from the measurement equation: 
$$y = \tau + \epsilon$$
and the prior for $\tau$ is 
$$\tau|\beta, c\sigma^2 \sim \mathcal{N}(H^{-1}\mu I_T, c\sigma^2(H^TH)^{-1})$$
Hence, we have 

$$y|\epsilon \sim \mathcal{N}(\epsilon + H^{-1}\mu I_T, c\sigma^2(H^TH)^{-1}) \propto exp\left(-\frac{1}{2}\frac{(y - \epsilon- H^{-1}\mu I_T)^TH^TH(y - \epsilon - H^{-1}\mu I_T)}{c\sigma^2}\right)$$
We have the prior for $\epsilon$: 
$$\epsilon | \sigma_e, \alpha \sim \mathcal{N}(0_T,\sigma_e^2(H_\alpha^TH_\alpha)^{-1}) \\ \propto exp\left(-\frac{1}{2\sigma^2_e}\epsilon^T(H_\alpha^TH_\alpha)\epsilon\right)$$
To derive the posterior distribution for $epsilon$, we use Baye's rule: 
$$\mathbf{p(\epsilon|y, \alpha, \beta, \sigma^2_e)} = \frac{\mathbf{p(\epsilon, y, \alpha, \beta, \sigma^2_e)}}{\mathbf{p(y, \alpha, \beta, \sigma^2_e)}} \propto \mathbf{p(\epsilon, y, \alpha, \beta, \sigma^2_e)} = \mathbf{p(y|\epsilon,  \alpha, \beta, \sigma^2_e)}p(\epsilon,  \alpha, \beta, \sigma^2_e)$$
$$= \mathbf{p(y|\epsilon, \beta)}p(\epsilon|\alpha, \beta, \sigma^2_e)p(\alpha, \beta, \sigma^2_e) \propto \mathbf{p(y|\epsilon, \beta)}p(\epsilon|\alpha, \sigma^2_e)$$
$$\propto exp\left(-\frac{1}{2}\frac{(y - \epsilon- H^{-1}\mu I_T)^TH^TH(y - \epsilon - H^{-1}\mu I_T)}{c\sigma^2}\right)exp\left(-\frac{1}{2\sigma^2_e}\epsilon^T(H_\alpha^TH_\alpha)\epsilon\right)$$

$$\propto exp\left( \frac{y^TH^THy - 2\epsilon^T H^T Hy - 2y^TH^THH^{-1}\mu I_T+\epsilon^T H^TH \epsilon + 2\epsilon^T H^THH^{-1}\mu I_T + I_T^T\mu (H^{-1})^TH^THH^{-1}\mu I_T
}{2c\sigma^2}\right)exp\left(\frac{1}{2\sigma^2_e}\epsilon^T(H_\alpha^TH_\alpha)\epsilon\right)$$

$$\propto exp\left(- \frac{c^{-1}\sigma^{-2}\epsilon^TH^THy + c^{-1}\sigma^{-2}\epsilon^TH^TH\epsilon+2c^{-1}\sigma^{-2}\mu\epsilon^TH^T + \sigma_e^{-2}\epsilon^T H_\alpha^TH_\alpha \epsilon }{2}\right)$$
$$= exp\left( \frac{\epsilon^T[c^{-1}\sigma^{-2}H^TH + \sigma_e^{-2}H_\alpha^T H_\alpha]\epsilon - 2\epsilon^T[c^{-1}\sigma^{-2}H^THy-c^{-1}\sigma^{-2}\mu H^T]}{2}\right)$$
Hence, $\epsilon|y, \alpha, \beta, \sigma^2_e \sim \mathcal{N}(\bar{\epsilon}, \bar{V}_\epsilon)$, where 

$$\bar{V}_\epsilon = [\color{purple}{\mathbf{c^{-1}\sigma^{-2}}}\color{black}{H^T H + \sigma_e^{-2}H_\alpha^T H_\alpha}]^{-1} \qquad \bar{\epsilon} = \bar{V}_\epsilon H^T H\color{purple}{\mathbf{c^{-1}\sigma^{-2}}}\color{black}{[y-\mu H^{-1}]}$$
3. $\mathbf{p(\alpha|y, \alpha, \epsilon, \sigma^2_e)}$
From state equation 2, we have: 
$$\epsilon = X_{\epsilon} \alpha + e \qquad \text{where} \qquad e \sim \mathcal{N}(0, \sigma^2_eI_T)$$
We note that $\alpha$ is not in the measurement equation, so the updates for $\alpha$ does not directly come from the data. Hence we have: 
$$p(\epsilon|\alpha , \sigma^2_e) \sim \mathcal{N}(X_\epsilon \alpha, \sigma^2_eI_T) \propto exp\left(\frac{(\epsilon - X_\epsilon\alpha)^T(\epsilon-X_\epsilon \alpha)}{2 \sigma^2_e}\right)$$
$$p(\alpha) \sim \mathcal{N}({\alpha_{prior}, V_{prior}})\mathbb{1}_{\alpha \in A} \propto exp\left({-\frac{1}{2}(\alpha - \alpha_{prior})^T}V_{\alpha,prior}^{-1}(\alpha - \alpha_{prior})\right)\mathbb{1}_{\alpha \in A} $$

Therefore, 

$$p(\alpha | y, \epsilon, \sigma_e^2) \propto p(\alpha, y, \epsilon, \sigma^2_e) \propto p(\epsilon|y, \sigma^2_e, \alpha)p(\alpha|y, \sigma^2_e)p(y, \sigma^2_e)\propto p(\epsilon|y, \sigma^2_e, \alpha)p(\alpha|y, \sigma^2_e)$$

$$\propto exp\left(\frac{(\epsilon - X_\epsilon\alpha)^T(\epsilon-X_\epsilon \alpha)}{2 \sigma^2_e}\right)exp\left({-\frac{1}{2}(\alpha - \alpha_{prior})^T}V_{\alpha,prior}^{-1}(\alpha - \alpha_{prior})\right)\mathbb{1}_{\alpha \in A}$$

$$\propto exp\left(\frac{-2\sigma^{-2}_e \alpha^TX_\epsilon^T\epsilon + \sigma^{-2}_e\alpha^TX_\epsilon X_\epsilon \alpha + \alpha^TV_{\alpha,prior}^{-1}\alpha - 2 \alpha^TV_{\alpha,prior}^{-1}\alpha_{prior}}{2} \right)\mathbb{1}_{\alpha \in A} $$
$$= exp\left(\frac{
\alpha^T(\sigma^{-2}_eX_\epsilon X_\epsilon + V_{\alpha,prior}^{-1}) \alpha - 2 \alpha^T(\sigma^{-2}_e X_\epsilon^T\epsilon + V_{\alpha,prior}^{-1}\alpha_{prior})}{2} \right)\mathbb{1}_{\alpha \in A}$$
Hence, we have the posterior distribution of $\alpha$ as: 

$$\alpha|y, \epsilon, \sigma^2_e \sim \mathcal{N}(\bar{\alpha}, \bar{V}_\alpha)\mathbb{1}_{\alpha \in A} \qquad where \qquad \bar{V}_\alpha = [\sigma^{-2}_eX_\epsilon^T X_\epsilon + V_{\alpha,prior}^{-1}]^{-1} \quad \bar{\alpha} = \bar{V}_\alpha[\sigma^{-2}_e X_\epsilon^T\epsilon + V_{\alpha,prior}^{-1}\alpha_{prior}]$$

4. $\mathbf{p(\beta|y, \tau)}$

From state equation 2, we have: $$H \tau = \ X_{\tau}\beta + \eta = \color{purple}{\mathbf{\mu I_T + \eta}}$$, 

where $\eta \sim \color{purple}{\mathbf{\mathcal{N}(0_T, c\sigma^2I_T)}}$ and $$H \tau \sim \color{purple}{\mathbf{\mathcal{N}(\mu I_T, c\sigma^2I_T)}} \propto exp\left(-\frac{(H\tau - \mu I_T)^T(H\tau - \mu I_T)}{2c\sigma^2}\right)$$

We have the prior of $\beta$ as: 
$$\beta = \mu\sim \mathcal{N}(\beta_{prior}, V_{\beta, prior}) \propto exp\left({-\frac{1}{2}(\beta - \beta_{prior})^T}V_{\beta,prior}^{-1}(\beta - \beta_{prior})\right) = exp\left({-\frac{1}{2}(\mu I_T - \beta_{prior})^T}V_{\beta,prior}^{-1}(\mu I_t - \beta_{prior})\right)$$

Then, the posterior distributuion of $\beta$ is:

$$p(\beta | y, \tau) \propto p(\beta, y, \tau) \propto p(H\tau | \beta,y)p(\beta) $$
$$\propto exp\left({-\frac{1}{2}(\mu I_T - \beta_{prior})^T}V_{\beta,prior}^{-1}(\mu I_t - \beta_{prior})\right)exp\left(-\frac{(H\tau - \mu I_T)^T(H\tau - \mu I_T)}{2c\sigma^2}\right)$$

$$\propto exp\left( \frac{\mu^2 V_{\beta,prior}^{-1} - 2\mu V_{\beta,prior}^{-1}\beta_{prior}-2\mu c^{-1}\sigma^{-2}H\tau+c^{-1}\sigma^{-2}\mu^2}{2}\right) \propto exp\left( \frac{\mu^2 V_{\beta,prior}^{-1} - 2\mu V_{\beta,prior}^{-1}\beta_{prior}-2\mu c^{-1}\sigma^{-2} H\tau+c^{-1}\sigma^{-2}\mu^2}{2}\right)$$
$$\propto  exp\left( \frac{\mu^2 (V_{\beta,prior}^{-1}+c^{-1}\sigma^{-2}) - 2\mu( V_{\beta,prior}^{-1}\beta_{prior} +c^{-1}\sigma^{-2} H\tau)}{2}\right) \sim \mathcal{N}(\bar{\beta}, \bar{V}_\beta)$$ and 

$$\bar{V}_\beta = (V_{\beta,prior}^{-1}+\color{purple}{\mathbf{c^{-1}\sigma^{-2}}}\color{black}{)^{-1}  \qquad \bar{\beta} = \bar{V}_\beta[V_{\beta,prior}^{-1}\beta_{prior}} +\color{purple}{\mathbf{c^{-1}\sigma^{-2}}\color{black}{ H\tau]}}$$


5. $\mathbf{p(\sigma^2_e|y, \epsilon, \alpha)}$ Again from state equation 2, we have: 

$$\epsilon = X_{\epsilon} \alpha + e \qquad \text{where} \qquad e \sim \mathcal{N}(0, \sigma^2_eI_T)$$
Hence, we have $$\epsilon | \alpha, \sigma^2_e \sim \mathcal{N}(X_\epsilon \alpha, \sigma_e^2) \propto (\sigma_e^2)^{\frac{T}{2}}exp(-\frac{(\epsilon - X_\epsilon \alpha)^T(\epsilon - X_\epsilon \alpha)}{2\sigma^2_e})$$

The prior distribution of $\sigma^2_e$ is assumed to follow an inverse-gamma 2 distribution, with 

$$\sigma^2_e \sim \mathcal{N}(s_{prior}, \nu_{prior}) \propto (\sigma^2_e)^{-\frac{\nu_{prior}+2}{2}}exp(-\frac{s_{prior}}{2\sigma^2_e})$$
Then, the posterior distribution of $\sigma^2_e$ can be derived as follows: 

$$p(\sigma^2_e|\epsilon, \alpha, \sigma^2_e, y) \propto \frac{p(\sigma^2_e, \epsilon, \alpha, \sigma^2_e, y)}{p(\epsilon, \alpha, \sigma^2_e, y)} \propto p(\sigma^2_e, \epsilon, \alpha, \sigma^2_e, y)$$ 
$$ \propto p(\epsilon|\sigma^2_e,\alpha, y)p(\sigma^2_e,\alpha, y)\propto p(\epsilon|\sigma^2_e,\alpha, y)p(\sigma^2_e)$$

$$\propto (\sigma^2_e)^{-\frac{\nu_{prior}+2}{2}}exp(-\frac{s_{prior}}{\sigma^2_e})(\sigma_e^2)^{\frac{T}{2}}exp(-\frac{(\epsilon - X_\epsilon \alpha)^T(\epsilon - X_\epsilon \alpha)}{2\sigma^2_e})$$
$$\propto (\sigma^2_e)^{-\frac{\nu_{prior}+T+2}{2}}exp(-\frac{(\epsilon - X_\epsilon \alpha)^T(\epsilon - X_\epsilon \alpha)+s_{prior}}{2\sigma^2_e}) \sim \mathcal{IG2}(\bar{s_{e}}, \bar{\nu_e})$$
where 
$$\bar{s}_e = s_{prior} + (\epsilon - X_\epsilon \alpha)^T(\epsilon - X_\epsilon \alpha) \qquad \bar{\nu}_e = T+ \nu_{prior}$$

Note that in this case, $\sigma^2_\eta$ is assumed to be a constant, with no posterior distribution. 

```{r}
UC.AR.Gibbs.sampler.Sigma.eta    = function(S, starting.values, priors, sigma_eta_fixed){
  # Estimation of a simple UC-AR(p) model with sigma_{\eta e}=0, sigma_eta fixed and tau_0 = 0
  
  aux     = starting.values
  p       = length(aux$alpha)
  T       = nrow(aux$Y)
  #####  #####  #####  #####  #####  #####  #####  #####  #####
  HH = t(aux$H)%*%aux$H
  #####  #####  #####  #####  #####  #####  #####  #####  #####
  posteriors    = list(
    tau     = matrix(NA,T,S),
    epsilon = matrix(NA,T,S),
    beta    = matrix(NA,2,S),
    alpha   = matrix(NA,p,S),
    sigma   = matrix(NA,2,S),
    non.stationary.iterations = rep(NA,S)
  )

  for (s in 1:S){

    # Sampling beta
    ###########################
    ######Modified from simple model####################################
    #beta.v.inv    = diag(1/diag(priors$beta.v))
    beta.v.inv = 1/priors$beta.v
    #V.beta.bar    = solve( (1/aux$sigma[1])*crossprod(priors$X.tau) + beta.v.inv )
    V.beta.bar    = (1/sigma_eta_fixed + beta.v.inv)^(-1)
    #beta.bar      = V.beta.bar %*% ( (1/aux$sigma[1])*crossprod(priors$X.tau, c(aux$tau[1],diff(aux$tau))) + beta.v.inv%*%priors$beta.m )
    beta.bar = V.beta.bar*(beta.v.inv*priors$beta.m + sigma_eta_fixed*aux$H*aux$tau)

    #beta.draw     = rmvnorm(1,as.vector(beta.bar),V.beta.bar)
    beta.draw     = rnorm(1,beta.bar,V.beta.bar)
    #aux$beta      = as.vector(beta.draw)
    aux$beta = beta.draw
    ####################################################
    X.epsilon     = matrix(NA,T,0)
    for (i in 1:p){
      X.epsilon   = cbind(X.epsilon, c(rep(0,i),aux$epsilon[1:(T-i),]))
    }

    # Sampling alpha
    ###########################
    alpha.v.inv   = diag(1/diag(priors$alpha.v))
    V.alpha.bar   = solve((1/aux$sigma[2])*crossprod(X.epsilon) + alpha.v.inv )
    V.alpha.bar   = 0.5*(V.alpha.bar + t(V.alpha.bar))
    alpha.bar     = V.alpha.bar %*% ( (1/aux$sigma[2])*crossprod(X.epsilon, aux$epsilon) + alpha.v.inv%*%priors$alpha.m )
    # non.stationary= TRUE
    # ns.i          = 1
    # while(non.stationary){
      alpha.draw    = rmvnorm(1,as.vector(alpha.bar),V.alpha.bar)
    #   non.stationary= max(Mod(eigen(rbind(alpha.draw,cbind(diag(p-1),rep(0,p-1))))$values))>=1
    #   ns.i          = ns.i + 1
    # }
    aux$alpha     = as.vector(alpha.draw)

    H.alpha       = diag(T)
    for (i in 1:p){
      sdiag(H.alpha,-i) =  -aux$alpha[i]
    }
    aux$H.alpha   = H.alpha

    # Sampling sigma
    ###########################
    #sigma.eta.s   = as.numeric(priors$sigma.s + crossprod( (c(aux$tau[1],diff(aux$tau)) - X.tau%*%aux$beta), (priors$H%*%aux$tau - X.tau%*%aux$beta)))
    #sigma.eta.nu  = priors$sigma.nu + T
    #sigma.eta.draw= sigma.eta.s/rchisq(1,sigma.eta.nu)

    sigma.e.s     = as.numeric(priors$sigma.s + crossprod(aux$H.alpha%*%aux$epsilon))
    sigma.e.nu    = priors$sigma.nu + T
    sigma.e.draw  = sigma.e.s/rchisq(1,sigma.e.nu)
    #aux$sigma     = c(sigma.eta.draw,sigma.e.draw)
    aux$sigma     = c(sigma_eta_fixed,sigma.e.draw)
    # Sampling tau
    ###########################
    #V.tau.inv     = (1/aux$sigma[2])*crossprod(aux$H.alpha) + (1/aux$sigma[1])*HH
    V.tau.inv     = (1/aux$sigma[2])*crossprod(aux$H.alpha) + (1/sigma_eta_fixed)*HH
    V.tau.inv     = 0.5*(V.tau.inv + t(V.tau.inv))
    #b.tau         = (1/aux$sigma[2])*crossprod(aux$H.alpha, aux$H.alpha%*%aux$Y) + (1/aux$sigma[1])*crossprod(H, X.tau%*%aux$beta)
    b.tau         = (1/aux$sigma[2])*crossprod(aux$H.alpha, aux$H.alpha%*%aux$Y) + (1/sigma_eta_fixed)*t(aux$H) * aux$beta
    precision.L   = t(bandchol(V.tau.inv))
    epsilon       = rnorm(T)
    b.tau.tmp     = forwardsolve(precision.L, b.tau)
    tau.draw      = backsolve(t(precision.L), b.tau.tmp + epsilon)
    aux$tau       = tau.draw

    # Sampling epsilon
    ###########################
    #b.epsilon     = (1/aux$sigma[1])*HH%*%(aux$Y- cumsum(X.tau%*%aux$beta))
    b.epsilon     = (1/sigma_eta_fixed)*HH%*%(aux$Y- aux$beta*solve(aux$H))
    epsilon.n     = rnorm(T)
    b.epsilon.tmp = forwardsolve(precision.L, b.epsilon)
    epsilon.draw  = backsolve(t(precision.L), b.epsilon.tmp + epsilon.n)
    aux$epsilon   = epsilon.draw

    posteriors$tau[,s]     = aux$tau
    posteriors$epsilon[,s] = aux$epsilon
    posteriors$beta[,s]    = aux$beta
    posteriors$alpha[,s]   = aux$alpha
    posteriors$sigma[,s]   = aux$sigma
    # posteriors$non.stationary.iterations[s] = ns.i
    if (s%%1000==0){cat(" ",s)}
  }

  output      = list(
    posterior = posteriors,
    last.draw = aux
  )
  return(output)
}
```
# Hierarchical modeling

## Estimating gamma error term variance prior scale

## Estimating inverted-gamma 2 error term variance prior scale

## Estimating the initial condition prior scale

## Student-t prior for the trend component

## Estimating Student-t degrees of freedom parameter

The Student-t distribution is commonly used in statistical modeling to handle data with heavier tails than the normal distribution. An essential parameter of the Student-t distribution is the degrees of freedom $\nu$, which controls the tail heaviness. In this note, we present the Bayesian estimation of the degrees of freedom parameter for an N-variate Student-t distribution using the Inverted-Gamma 2 (IG2) scale mixture of normals.

The N-variate Student-t distribution can be represented as a scale mixture of normals:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

where:

- $\mathbf{y}$ is the $N$-dimensional observation vector.
- $\mathbf{\mu}$ is the mean vector.
- $\lambda$ is the latent scale variable.
- $\nu$ is the degrees of freedom parameter.

### Derivation of Full Conditional Posteriors

#### Full Conditional Posterior of $\lambda$

Given the prior distribution:

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

The likelihood of the data given $\lambda$ is:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

The full conditional posterior of $\lambda$ can be derived as follows:

1. **Likelihood of $\mathbf{y}$ given $\mathbf{\mu}$ and $\lambda$**:

   $$
   p(\mathbf{y} \mid \mathbf{\mu}, \lambda) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right)
   $$

2. **Prior for $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) \propto \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

3. **Joint distribution**:

   $$
   p(\mathbf{y}, \lambda \mid \mathbf{\mu}, \nu) = p(\mathbf{y} \mid \mathbf{\mu}, \lambda) p(\lambda \mid \nu)
   $$

4. **Full conditional posterior**:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right) \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

   Combining terms:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\left(\nu + \frac{N}{2} + 1\right)} \exp\left(-\frac{\nu + \frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2}}{\lambda}\right)
   $$

   This is recognized as the kernel of an Inverted-Gamma 2 distribution:

   $$
   \lambda \mid \mathbf{y}, \mathbf{\mu}, \nu \sim \mathcal{IG2}\left(\nu + N, \nu + (\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})\right)
   $$

#### Full Conditional Posterior of $\nu$

To estimate $\nu$, we use the Metropolis-Hastings algorithm due to its non-standard form. The steps for deriving the full conditional posterior of $\nu$ are as follows:

1. **Likelihood of $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) = \frac{\left(\frac{\nu}{2}\right)^{\nu/2}}{\Gamma(\nu/2)} \lambda^{-\left(\nu/2 + 1\right)} \exp\left(-\frac{\nu}{2\lambda}\right)
   $$

2. **Log-likelihood for $\nu$ given $\lambda$**:

   $$
   \log p(\lambda \mid \nu) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

3. **Log-prior for $\nu$ (assuming a non-informative prior)**:

   $$
   \log p(\nu) = \text{constant}
   $$

4. **Full conditional posterior**:

   The full conditional posterior for $\nu$ is proportional to the product of the likelihood and the prior:

   $$
   p(\nu \mid \lambda) \propto p(\lambda \mid \nu) p(\nu)
   $$

   Since $p(\nu)$ is constant, we focus on $p(\lambda \mid \nu)$:

   $$
   \log p(\nu \mid \lambda) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

   This expression does not have a closed form, so we use the Metropolis-Hastings algorithm to sample from this posterior.

### R Function for Gibbs Sampler

Below is the R function implementing the Gibbs sampler for estimating $\nu$ using the IG2-scale mixture of normals representation.

```{r echo=TRUE}
metropolis_hastings_nu <- function(y, mu, n_iter, init_nu, proposal_sd) {
  # Initialize parameter
  nu <- init_nu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  
  # Log-likelihood function
  log_likelihood <- function(nu, y, mu) {
    sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
  }
  
  for (i in 1:n_iter) {
    # Propose new value for nu
    nu_proposal <- nu + rnorm(1, 0, proposal_sd)
    
    if (nu_proposal > 0) {
      # Calculate log acceptance ratio
      log_acceptance_ratio <- log_likelihood(nu_proposal, y, mu) - log_likelihood(nu, y, mu)
      
      # Accept or reject the proposal
      if (log(runif(1)) < log_acceptance_ratio) {
        nu <- nu_proposal
      }
    }
    
    # Store the sample
    nu_samples[i] <- nu
  }
  
  return(nu_samples)
}

gibbs_sampler_t <- function(y, n_iter, init_values) {
  # Initialize parameters
  nu <- init_values$nu
  mu <- init_values$mu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  mu_samples <- numeric(n_iter)
  lambda_samples <- numeric(n_iter)
  
  for (i in 1:n_iter) {
    # Sample lambda
    shape_lambda <- nu + N
    rate_lambda <- nu + sum((y - mu)^2)
    lambda <- 1 / rgamma(1, shape = shape_lambda, rate = rate_lambda)
    
    # Sample mu
    mu <- rnorm(1, mean = mean(y), sd = sqrt(lambda / N))
    
    # Sample nu using Metropolis-Hastings
    log_likelihood <- function(nu, y, mu) {
      sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
    }
    
    proposal_nu <- nu + rnorm(1, 0, 0.1) # proposal distribution: normal random walk
    if (proposal_nu > 0) {
      log_acceptance_ratio <- log_likelihood(proposal_nu, y, mu) - log_likelihood(nu, y, mu)
      if (log(runif(1)) < log_acceptance_ratio

) {
        nu <- proposal_nu
      }
    }
    
    # Store samples
    nu_samples[i] <- nu
    mu_samples[i] <- mu
    lambda_samples[i] <- lambda
  }
  
  return(list(nu = nu_samples, mu = mu_samples, lambda = lambda_samples))
}

# Example usage
set.seed(123)
y <- rnorm(100)
init_values <- list(nu = 5, mu = mean(y))
n_iter <- 1000
result <- gibbs_sampler_t(y, n_iter, init_values)

# Display the results
print(summary(result$nu))
print(summary(result$mu))
print(summary(result$lambda))
```

### Conclusion

This note provided a comprehensive step-by-step algebraic derivation and a sampler for estimating the degrees of freedom parameter $\nu$ for an N-variate Student-t distribution using an IG2-scale mixture of normals approach within a Bayesian framework. By using the Metropolis-Hastings algorithm, we avoid the need to assume a prior distribution for $\nu$, simplifying the estimation process. This approach allows for flexible modeling of heavy-tailed data, which is often encountered in practice.

### References

- Geweke, J. (1993). Bayesian treatment of the independent Student-t linear model. *Journal of Applied Econometrics*, 8(S1), S19-S40.
- Chib, S., & Greenberg, E. (1995). Understanding the Metropolis-Hastings Algorithm. *The American Statistician*, 49(4), 327-335.




## Laplace prior for the trend component

# Model extensions

## Autoregressive cycle component

## Random walk with  time-varying drift parameter

## Student-t error terms

## Conditional heteroskedasticity

# Bayesian forecasting

## Predictive density

## Sampling from the predictive density

## Missing observations


## References {.unnumbered}
