---
title: "Bayesian Unobserved Component Models"
author:
  - name: Tomasz Woźniak
    url: https://github.com/donotdespair
    orcid: 0000-0003-2212-2378
  - name: Ben Gussen
    url: https://github.com/bgussen
    orcid: 0000-0003-4406-4076
  - name: Zheyuan Li
    url: https://github.com/lzyzero3
  - name: Qingqing Pang
    url: https://github.com/comeseemeintheocean/Macro-template
  
    execute:
  #echo: false
citation: 
  issued: 2024-05-01
  url: https://donotdespair.github.io/Bayesian-Unobserved-Component-Models/
  doi: 10.26188/25814617
bibliography: references.bib
---

> **Abstract.** We present the basics of Bayesian estimation and inference for unobserved component models on the example of a local-level model. 
> The range of topics includes the conjugate prior analysis using normal-inverted-gamma 2 distribution and its extensions focusing on hierarchical modelling, conditional heteroskedasticity, and Student-t error terms. 
> We scrutinise Bayesian forecasting and sampling from the predictive density.
>
> **Keywords.** Unobserved Component Models, Local-Level Model, State-Space Bayesian Inference, Forecasting, Heteroskedasticity, Hierarchical Modelling, Gibbs Sampler, Simulation Smoother, Precision Sampling

# Unobserved component models

Unobserved Component (UC) models are a popular class of models in macroeconometrics that use the state-space representation for unit-root nonstationary time series. 
The simple formulation of the model equations decomposing the series into a non-stationary and stationary component facilitates economic interpretations and good forecasting performance.

# A simple local-level model

The model is set for a univariate time series whose observation at time
$t$ is denoted by $y_t$. 
It decomposes the variable into a stochastic trend component, $\tau_t$, and a stationary error component, $\epsilon_t$. 
The former follows a Gaussian random walk process with the conditional variance $\sigma_\eta^2$, and the latter is zero-mean normally distributed with the variance $\sigma^2$.
These are expressed as the model equations:

$$
\begin{align}
y_t &= \tau_t + \epsilon_t,\\
\tau_t &= \tau_{t-1} + \eta_t,\\
\epsilon_t &\sim\mathcal{N}\left(0, \sigma^2\right),\\
\eta_t &\sim\mathcal{N}\left(0, \sigma_\eta^2\right),
\end{align}
$$
where the initial condition $\tau_0$ is a parameter of the model.

## Matrix notation for the model

To simplify the notation and the derivations introduce matrix notation for the model. Let $T$ be the available sample size for the variable $y$. 
Define a $T$-vector of zeros, $\mathbf{0}_T$, and of ones, $\boldsymbol\imath_T$, the identity matrix of order $T$, $\mathbf{I}_T$, as well as $T\times1$ vectors:

$$
\begin{align}
\mathbf{y} = \begin{bmatrix} y_1\\ \vdots\\ y_T \end{bmatrix},\quad
\boldsymbol\tau = \begin{bmatrix} \tau_1\\ \vdots\\ \tau_T \end{bmatrix},\quad
\boldsymbol\epsilon = \begin{bmatrix} \epsilon_1\\ \vdots\\ \epsilon_T \end{bmatrix},\quad
\boldsymbol\eta = \begin{bmatrix} \eta_1\\ \vdots\\ \eta_T \end{bmatrix},\qquad
\mathbf{i} = \begin{bmatrix} 1\\0\\ \vdots\\ 0 \end{bmatrix},
\end{align}
$$
and a $T\times T$ matrix $\mathbf{H}$ with the elements:

$$
\begin{align}
\mathbf{H} = \begin{bmatrix}
1 & 0 & \cdots & 0 & 0\\
-1 & 1 & \cdots & 0 & 0\\
0 & -1 & \cdots & 0 & 0\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
0 & 0 & \cdots & 1 & 0\\
0 & 0 & \cdots & -1 & 1
\end{bmatrix}.
\end{align}
$$


Then the model can be written in a concise notation as:
$$
\begin{align}
\mathbf{y} &= \mathbf{\tau} + \boldsymbol\epsilon,\\
\mathbf{H}\boldsymbol\tau &= \mathbf{i} \tau_0 + \boldsymbol\eta,\\
\boldsymbol\epsilon &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma^2\mathbf{I}_T\right),\\
\boldsymbol\eta &\sim\mathcal{N}\left(\mathbf{0}_T, \sigma_\eta^2\mathbf{I}_T\right).
\end{align}
$$

## Likelihood function

The model equations imply the predictive density of the data vector
$\mathbf{y}$. To see this, consider the model equation as a linear
transformation of a normal vector $\boldsymbol\epsilon$. Therefore, the data vector follows a multivariate normal distribution given by:
$$
\begin{align}
\mathbf{y}\mid \boldsymbol\tau, \sigma^2 &\sim\mathcal{N}_T\left(\boldsymbol\tau, \sigma^2\mathbf{I}_T\right).
\end{align}
$$

This distribution determines the shape of the likelihood function that
is defined as the sampling data density: 
$$
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y})\equiv p\left(\mathbf{y}\mid\boldsymbol\tau, \sigma^2 \right).
\end{align}
$$

The likelihood function that for the sake of the estimation of the parameters, and after plugging in data in place of $\mathbf{y}$, is considered a function of parameters $\boldsymbol\tau$ and $\sigma^2$ is given by: 
$$
\begin{align}
L(\boldsymbol\tau,\sigma^2|\mathbf{y}) = 
(2\pi)^{-\frac{T}{2}}\left(\sigma^2\right)^{-\frac{T}{2}}\exp\left\{-\frac{1}{2}\frac{1}{\sigma^2}(\mathbf{y} - \boldsymbol\tau)'(\mathbf{y} - \boldsymbol\tau)\right\}.
\end{align}
$$

## Prior distributions

# Bayesian estimation

## Gibbs sampler

## Simulation smoother and precision sampler

## Analytical solution for a joint posterior

# Hierarchical modeling

## Estimating gamma error term variance prior scale
To estimate the gamma error term variance, we need to firstly put a prior on the prior scale of and present its full conditional posterior distribution.

The prior is below:

```{=tex}
\begin{align*}
\sigma^2 \mid s &\sim \text{IG2}(s, \nu) \\
s &\sim \mathcal{G}(s, a)
\end{align*}
```


Then we use Gibbs sampling to draw samples from the full conditional distributions: 
```{r}
#| echo: true
#| message: false
#| warning: false
library(MCMCpack)
# Define prior parameters
s_prior <- 2
nu_prior <- 2
a_prior <- 2

# Define the function for the conditional posterior distribution
posterior_sampler <- function(y, n_iter = 1000) {
  # Store samples
  s <- rgamma(1, shape = a_prior, rate = s_prior)
  sigma2 <- rinvgamma(1, shape = s, scale = nu_prior)  

  s_samples <- numeric(n_iter)
  sigma2_samples <- numeric(n_iter)

  for (i in 1:n_iter) {
    shape_sigma2 <- (length(y) / 2) + s
    rate_sigma2 <- (sum((y - mean(y))^2) / 2) + nu_prior
    sigma2 <- rinvgamma(1, shape = shape_sigma2, scale = rate_sigma2)  

    shape_s <- a_prior + 1
    rate_s <- s_prior + sigma2
    s <- rgamma(1, shape = shape_s, rate = rate_s)

    sigma2_samples[i] <- sigma2
    s_samples[i] <- s
  }

  return(list(sigma2_samples = sigma2_samples, s_samples = s_samples))
}

set.seed(123)
y <- rnorm(100, mean = 0, sd = 1)

# Run the sampler
posterior_samples <- posterior_sampler(y, n_iter = 1000)

# Plot posterior distributions
par(mfrow = c(2, 1))
hist(posterior_samples$sigma2_samples, main = expression(paste("Posterior of ", sigma^2)), xlab = expression(sigma^2))
hist(posterior_samples$s_samples, main = "Posterior of s", xlab = "s")
```

Based on the sampling results, we summarize the posterior distribution of these parameters.

As can be seen from the histogram, the values of $\sigma^2$ are mostly concentrated between 0.6 and 1.0. The distribution has a certain skewness, with a small number of higher values on the right, but it is generally concentrated.

Most of the values of $s$ are concentrated between 0 and 2, indicating that the probability of s is higher in this range and the distribution has a longer right tail.

To derive the full conditional posteriors for $\sigma^2$ and $s$, we start with the joint posterior distribution of these parameters given the data $y$.

The joint posterior distribution is below:

```{=tex}
\begin{equation}
p(\sigma^2, s \mid y) \propto p(y \mid \sigma^2) p(\sigma^2 \mid s) p(s)
\end{equation}
```


Full Conditional Posterior for $\sigma^2$ is: 

```{=tex}
\begin{equation}
p(\sigma^2 \mid y, s) \propto (\sigma^2)^{-\left( \frac{T}{2} + s + 1 \right)} \exp \left( -\frac{1}{\sigma^2} \left( \frac{(y - \tau)'(y - \tau)}{2} + \nu \right) \right)
\end{equation}
```


Full Conditional Posterior for $s$ is:

```{=tex}
\begin{equation}
p(s \mid \sigma^2) \propto s^{a-1} \nu^s (\sigma^2)^{-(s+1)} \exp \left( -s - \frac{\nu}{\sigma^2} \right)
\end{equation}
```


## Estimating inverted-gamma 2 error term variance prior scale

## Estimating the initial condition prior scale

## Student-t prior for the trend component

## Estimating Student-t degrees of freedom parameter

The Student-t distribution is commonly used in statistical modeling to handle data with heavier tails than the normal distribution. An essential parameter of the Student-t distribution is the degrees of freedom $\nu$, which controls the tail heaviness. In this note, we present the Bayesian estimation of the degrees of freedom parameter for an N-variate Student-t distribution using the Inverted-Gamma 2 (IG2) scale mixture of normals.

The N-variate Student-t distribution can be represented as a scale mixture of normals:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

where:

- $\mathbf{y}$ is the $N$-dimensional observation vector.
- $\mathbf{\mu}$ is the mean vector.
- $\lambda$ is the latent scale variable.
- $\nu$ is the degrees of freedom parameter.

### Derivation of Full Conditional Posteriors

#### Full Conditional Posterior of $\lambda$

Given the prior distribution:

$$
\lambda \mid \nu \sim \mathcal{IG2}(\nu, \nu)
$$

The likelihood of the data given $\lambda$ is:

$$
\mathbf{y} \mid \mathbf{\mu}, \lambda \sim \mathcal{N}(\mathbf{\mu}, \lambda \mathbf{I}_N)
$$

The full conditional posterior of $\lambda$ can be derived as follows:

1. **Likelihood of $\mathbf{y}$ given $\mathbf{\mu}$ and $\lambda$**:

   $$
   p(\mathbf{y} \mid \mathbf{\mu}, \lambda) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right)
   $$

2. **Prior for $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) \propto \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

3. **Joint distribution**:

   $$
   p(\mathbf{y}, \lambda \mid \mathbf{\mu}, \nu) = p(\mathbf{y} \mid \mathbf{\mu}, \lambda) p(\lambda \mid \nu)
   $$

4. **Full conditional posterior**:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\frac{N}{2}} \exp\left(-\frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2\lambda}\right) \lambda^{-\nu - 1} \exp\left(-\frac{\nu}{\lambda}\right)
   $$

   Combining terms:

   $$
   p(\lambda \mid \mathbf{y}, \mathbf{\mu}, \nu) \propto \lambda^{-\left(\nu + \frac{N}{2} + 1\right)} \exp\left(-\frac{\nu + \frac{(\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})}{2}}{\lambda}\right)
   $$

   This is recognized as the kernel of an Inverted-Gamma 2 distribution:

   $$
   \lambda \mid \mathbf{y}, \mathbf{\mu}, \nu \sim \mathcal{IG2}\left(\nu + N, \nu + (\mathbf{y} - \mathbf{\mu})'(\mathbf{y} - \mathbf{\mu})\right)
   $$

#### Full Conditional Posterior of $\nu$

To estimate $\nu$, we use the Metropolis-Hastings algorithm due to its non-standard form. The steps for deriving the full conditional posterior of $\nu$ are as follows:

1. **Likelihood of $\lambda$ given $\nu$**:

   $$
   p(\lambda \mid \nu) = \frac{\left(\frac{\nu}{2}\right)^{\nu/2}}{\Gamma(\nu/2)} \lambda^{-\left(\nu/2 + 1\right)} \exp\left(-\frac{\nu}{2\lambda}\right)
   $$

2. **Log-likelihood for $\nu$ given $\lambda$**:

   $$
   \log p(\lambda \mid \nu) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

3. **Log-prior for $\nu$ (assuming a non-informative prior)**:

   $$
   \log p(\nu) = \text{constant}
   $$

4. **Full conditional posterior**:

   The full conditional posterior for $\nu$ is proportional to the product of the likelihood and the prior:

   $$
   p(\nu \mid \lambda) \propto p(\lambda \mid \nu) p(\nu)
   $$

   Since $p(\nu)$ is constant, we focus on $p(\lambda \mid \nu)$:

   $$
   \log p(\nu \mid \lambda) = \frac{\nu}{2} \log\left(\frac{\nu}{2}\right) - \log\Gamma\left(\frac{\nu}{2}\right) - \left(\frac{\nu}{2} + 1\right) \log \lambda - \frac{\nu}{2\lambda}
   $$

   This expression does not have a closed form, so we use the Metropolis-Hastings algorithm to sample from this posterior.

### R Function for Gibbs Sampler

Below is the R function implementing the Gibbs sampler for estimating $\nu$ using the IG2-scale mixture of normals representation.

```{r echo=TRUE}
metropolis_hastings_nu <- function(y, mu, n_iter, init_nu, proposal_sd) {
  # Initialize parameter
  nu <- init_nu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  
  # Log-likelihood function
  log_likelihood <- function(nu, y, mu) {
    sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
  }
  
  for (i in 1:n_iter) {
    # Propose new value for nu
    nu_proposal <- nu + rnorm(1, 0, proposal_sd)
    
    if (nu_proposal > 0) {
      # Calculate log acceptance ratio
      log_acceptance_ratio <- log_likelihood(nu_proposal, y, mu) - log_likelihood(nu, y, mu)
      
      # Accept or reject the proposal
      if (log(runif(1)) < log_acceptance_ratio) {
        nu <- nu_proposal
      }
    }
    
    # Store the sample
    nu_samples[i] <- nu
  }
  
  return(nu_samples)
}

gibbs_sampler_t <- function(y, n_iter, init_values) {
  # Initialize parameters
  nu <- init_values$nu
  mu <- init_values$mu
  N <- length(y)
  
  # Storage for samples
  nu_samples <- numeric(n_iter)
  mu_samples <- numeric(n_iter)
  lambda_samples <- numeric(n_iter)
  
  for (i in 1:n_iter) {
    # Sample lambda
    shape_lambda <- nu + N
    rate_lambda <- nu + sum((y - mu)^2)
    lambda <- 1 / rgamma(1, shape = shape_lambda, rate = rate_lambda)
    
    # Sample mu
    mu <- rnorm(1, mean = mean(y), sd = sqrt(lambda / N))
    
    # Sample nu using Metropolis-Hastings
    log_likelihood <- function(nu, y, mu) {
      sum(dt((y - mu) / sqrt(nu), df = nu, log = TRUE))
    }
    
    proposal_nu <- nu + rnorm(1, 0, 0.1) # proposal distribution: normal random walk
    if (proposal_nu > 0) {
      log_acceptance_ratio <- log_likelihood(proposal_nu, y, mu) - log_likelihood(nu, y, mu)
      if (log(runif(1)) < log_acceptance_ratio

) {
        nu <- proposal_nu
      }
    }
    
    # Store samples
    nu_samples[i] <- nu
    mu_samples[i] <- mu
    lambda_samples[i] <- lambda
  }
  
  return(list(nu = nu_samples, mu = mu_samples, lambda = lambda_samples))
}

# Example usage
set.seed(123)
y <- rnorm(100)
init_values <- list(nu = 5, mu = mean(y))
n_iter <- 1000
result <- gibbs_sampler_t(y, n_iter, init_values)

# Display the results
print(summary(result$nu))
print(summary(result$mu))
print(summary(result$lambda))
```

### Conclusion

This note provided a comprehensive step-by-step algebraic derivation and a sampler for estimating the degrees of freedom parameter $\nu$ for an N-variate Student-t distribution using an IG2-scale mixture of normals approach within a Bayesian framework. By using the Metropolis-Hastings algorithm, we avoid the need to assume a prior distribution for $\nu$, simplifying the estimation process. This approach allows for flexible modeling of heavy-tailed data, which is often encountered in practice.

### References

- Geweke, J. (1993). Bayesian treatment of the independent Student-t linear model. *Journal of Applied Econometrics*, 8(S1), S19-S40.
- Chib, S., & Greenberg, E. (1995). Understanding the Metropolis-Hastings Algorithm. *The American Statistician*, 49(4), 327-335.


## Laplace prior for the trend component
To apply the Laplace prior for our trend component, we have $\lambda_t$ following an exponential distribution with mean alpha, where $\lambda_t \sim exp({\frac{1}{\alpha}})$, with the prior of the trend component is $\tau|\eta,\lambda \sim \mathcal{N}(H^{-1}i\tau_0, \sigma_\eta^2 H^{-1}\Omega(H^{-1})')$, then we have the marginal distribution of $\tau$ is Laplace distribution.

The model is where $\Omega=diag(\lambda_1,\lambda_2,...,\lambda_t)$, each lambda independently drawn from an exponential distribution.

Due to the change prior distribution of the trend component, now we have different full posterior distributions for parameters and in this note we only focus on the derivation for $\tau$ and $\lambda_t$.
The change of the prior distribution for $\tau$ gives the followings:

$$\tau =y - \epsilon$$

$$\tau = H^{-1}i\tau_0 + H^{-1} \eta $$

$$\tau|\eta,\lambda \sim \mathcal{N}(H^{-1}i\tau_0, \sigma_\eta^2 H^{-1}\Omega(H^{-1})') $$

$${\eta|\lambda \sim N(0_T, \sigma_\eta^2\Omega)}$$

$$\lambda_t \sim exp({\frac{1}{\alpha}})$$

### Full-conditional posterior distribution for $\tau|\tau_0,\sigma^2,\Omega$

**Conditional Likelihood:** 
$$\tau =y - \epsilon$$

$$\epsilon \sim \mathcal{N}(0_T, \sigma^2I_T)$$

$$L(\tau|y,\sigma^2) = exp({-\frac{1}{2}\frac{1}{\sigma^2}(\mathbf{y} - \boldsymbol\tau)'(\mathbf{y} - \boldsymbol\tau)})$$

The **prior distribution** of $\tau$is:

$$p(\tau|\tau_0,\sigma_\eta^2,\Omega) \propto exp(-\frac{1}{2}\frac{1}{\sigma_\eta^2}(\tau - H^{-1}i\tau_0)'H'\Omega^{-1}H(\tau - H^{-1}i\tau_0))$$
**Full-conditional posterior distribution** of ${\tau}|y,\sigma_\eta^2,\Omega$:
```{=tex}
\begin{align}
P(\tau|y,\tau_0,\sigma^2,\Omega) &\propto L(\tau|y,\sigma^2)\times p(\tau|\tau_0,\sigma_\eta^2,\Omega)\\

&\propto exp({-\frac{1}{2}}\sigma^{-2}\ \times (\tau'\tau -2y'\tau+y'y))\\
&\times exp({-\frac{1}{2}\sigma_\eta^{-2}\ (\tau'H'\Omega^{-1}H\tau - 2\tau'H'\Omega^{-1}H(H^{-1}i\tau_0) +(H^{-1}i\tau_0)'H\Omega^{-1}H(H^{-1}i\tau_0)}\\

&\propto exp({-\frac{1}{2}}\sigma^{-2}\sigma_\eta^{-2}[\tau'(\sigma_\eta^2+\sigma^2H'\Omega^{-1}H)\tau-2\tau'(\sigma_\eta^2y+\sigma^2H'\Omega^{-1}H(H^{-1}i\tau_0))])\\

&= exp({-\frac{1}{2}}(\tau'(\sigma^{-2}+\sigma_\eta^{-2}H'\Omega^{-1}H\tau-2\tau'(\sigma^{-2}y+\sigma_\eta^{-2}H'\Omega^{-1}H(H^{-1}i\tau_0))\\

&\sim N(\bar{V_\tau},\bar{\tau})\\

&\bar{V_\tau} = (\sigma^{-2}+\sigma_\eta^{-2}H'\Omega^{-1}H)^{-1}\\
&\bar{\tau} =\bar{V_\tau} (\sigma^{-2}y+\sigma_\eta^{-2}H'\Omega^{-1}i\tau_0)

\end{align}
```


### Full-conditional posterior distribution for $\lambda_t|y,\tau,\sigma_\eta^2$

**Conditional Likelihood:** 

$$H\tau - i\tau_o=\eta$$
```{=tex}
\begin{align}
L(\lambda_t|y,\tau_0,\tau,\sigma_\eta^2)&\propto  (\prod^{T}_{i = 1} \lambda_t^{-\frac{N}{2}}) exp({-\frac{1}{2}\frac{1}{\sigma^2_\eta}(i\tau_0-H\tau)'\Omega^{-1}(i\tau_0-H\tau)})\\
&= (\prod^{T}_{i = 1} \lambda_t^{-\frac{N}{2}}exp({-\frac{1}{2}\frac{1}{\sigma^2_\eta}\frac{1}{\lambda_t}\eta_t'\eta_t}))                

\end{align}
```

The **prior distribution** of $\lambda_t$ is:

$$
P(\lambda_t) ={\frac{1}{\alpha}}exp({{-\frac{1}{\alpha}}\lambda_t)}) 
$$


**Full-conditional posterior distribution** of $\lambda_t$is:

```{=tex}
\begin{align}
P(\lambda_t|y,\tau,\sigma_\eta^2) &\propto \lambda_t^{-\frac{N}{2}+1-1}exp(-\frac{1}{2}({\frac{\eta_t'\sigma_\eta^{-2}\eta_t}{\lambda_t}+{\frac{2}{\alpha}}\lambda_t)}) \\
\end{align}
```


The above expression can be rearranged in the form of a Generalized inverse Gaussian distribution kernel as follows:

```{=tex}
\begin{align}
\lambda_t|Y,A,\Sigma &\sim GIG(a,b,p) \\
\\
a &=\frac{2}{\alpha} \\
b &= \eta_t'\sigma_\eta^{-2}\eta_t \\
p &= -\frac{N}{2}+1
\end{align}

```

### R Function for Gibbs Sample

```{r}
UC.AR.Gibbs.sampler    = function(S, starting.values, priors){
  aux     = starting.values
  p       = length(aux$alpha)
  T       = nrow(aux$Y)
#  i_matrix <- diag(T)
  i <- matrix(0, T, 1)  
  i[1, 1] <- 1 
  posteriors    = list(
    tau     = matrix(NA,T,S),
    epsilon = matrix(NA,T,S),
    tau_0   = matrix(NA,1,S),
    sigma   = matrix(NA,2,S),
    
    non.stationary.iterations = rep(NA,S)
  )
  
  alpha <- 2
  lambda.0 <- rexp(T, rate = 1/alpha)
  lambda.priors = list(alpha = 2)
  lambda.posterior.draws = array(NA,c(T,S+1))
  
  for (s in 1:S){
    
    if (s == 1) {
      lambda.s = lambda.0
    } else {
      lambda.s    = lambda.posterior.draws[,s]
    }
    
    Omega = (diag(lambda.s))
    Omega.inv = diag(1/lambda.s)
    
    # Sampling tau
    ###########################
    V.tau.inv     = (1/aux$sigma[2]) + (1/aux$sigma[1])*t(H) %*% Omega.inv %*% H
    V.tau.inv     = 0.5*(V.tau.inv + t(V.tau.inv))
    b.tau         = (1/aux$sigma[2])%*%aux$Y + (1/aux$sigma[1])*crossprod(H, Omega.inv%*%i%*%priors$tau_0)
    precision.L   = t(bandchol(V.tau.inv))
    epsilon       = rnorm(T)
    b.tau.tmp     = forwardsolve(precision.L, b.tau)
    tau.draw      = backsolve(t(precision.L), b.tau.tmp + epsilon)
    aux$tau       = tau.draw
    
    # Sampling tau_0
    ###########################
    tau_0.v.inv    = diag(1/diag(priors$tau_0.v))
    V.tau_0.bar    = solve((1/aux$sigma[1])*crossprod(i,Omega.inv%*%i) + tau_0.v.inv)
    tau_0.bar      = V.tau_0.bar %*% ( (1/aux$sigma[1])%*%t(i)%*%Omega.inv%*%H%*%aux$tau + tau_0.v.inv%*%priors$tau_0.m )
    tau_0.draw     = rmvnorm(1,as.vector(tau_0.bar),V.tau_0.bar)
    aux$tau_0      = as.vector(tau_0.draw)
    
    # Sampling sigma_eta
    ###########################
    sigma.eta.s   = as.numeric(priors$sigma.s + crossprod((c(aux$tau_0[1],diff(aux$tau_0)) - i%*%aux$tau_0),(priors$H%*%aux$tau - i%*%aux$tau_0)))
    sigma.eta.nu  = priors$sigma.nu + T
    sigma.eta.draw= sigma.eta.s/rchisq(1,sigma.eta.nu)
    sigma.eta.draw.inv=solve(sigma.eta.draw)
    
    sigma.e.s     = as.numeric(priors$sigma.s + crossprod(aux$tau-aux$Y))
    sigma.e.nu    = priors$sigma.nu + T
    sigma.e.draw  = sigma.e.s/rchisq(1,sigma.e.nu)
    aux$sigma     = c(sigma.eta.draw,sigma.e.draw)

    # Sampling lambda
    ###########################
    u.t = H%*%aux$tau[,,s]-i%*%aux$tau_0[,,s]
    #    ---- loop lambda posterior ----   #
    c                      = -N/2 + 1         
    a                      = 2 / lambda.priors$alpha
    for (x in 1:T){
      b                  = t((u.t)[x,])%*%(1/aux$sigma[1][,,s])%*%(u.t)[x,]
      lambda.posterior.draws[x,s+1] = GIGrvg::rgig(1, lambda = c, chi = b, psi = a)
    }

    aux$lambda = lambda.posterior.draws
  
    
    posteriors$tau[,s]     = aux$tau
    posteriors$tau_0[,s]   = aux$tau_0
    posteriors$lambda[,s]  = aux$lambda
    # posteriors$non.stationary.iterations[s] = ns.i
    if (s%%1000==0){cat(" ",s)}
  }
  
  output      = list(
    posterior = posteriors,
    last.draw = aux
  )
  return(output)
}

```



# Model extensions

## Autoregressive cycle component

## Random walk with  time-varying drift parameter

## Student-t error terms

## Conditional heteroskedasticity

# Bayesian forecasting

## Predictive density

## Sampling from the predictive density

## Missing observations


## References {.unnumbered}
